---
title: "IDS investigation worksheet"
author: Xu Sun
date: "`r Sys.Date()`"
output: html_document
---

**Note:** You can use this file as you 'working document' where you can try out various investigation ideas and keep notes about your findings. How you use and structure this file is up to you. It is recommended that you keep notes about what you are investigating and what you find as this will make the process of creating your presentation and report easier. Please note that you _do not_ need to submit this file as part of your group project.



```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r load-lib, message = FALSE}
library(tidyverse)
library(ggplot2)
# Add any other libraries here



```


```{r load-data}
# Load my data 
communities <- read_csv("data/communities.data")
glimpse(communities)

# Remove identifiers & Check missing values
data <- communities %>%
  select(-c(state, county, community, communityname, fold))
colSums(is.na(data)) %>% sort(decreasing = TRUE)
data_clean <- data %>% select(where(~ mean(is.na(.)) < 0.3))
data_clean <- data_clean %>% mutate(across(everything(), ~ replace_na(., mean(., na.rm = TRUE))))

str(data_clean)
summary(data_clean$ViolentCrimesPerPop)

# Distribution of Violent Crime
ggplot(data = data_clean, mapping = aes(x = ViolentCrimesPerPop)) +
  geom_histogram(bins = 50, fill = "navyblue", color = "white") +
  labs(title = "Distribution of Violent Crimes per Population", x = "Violent Crime Rate", y = "Count")

# Education vs. Crime
ggplot(data = data_clean, mapping = aes(x = PctNotHSGrad, y = ViolentCrimesPerPop)) +
  geom_point(alpha = 0.3) +
  geom_smooth(method = "lm", color = "yellow") +
  labs(title = "Violent Crime vs. % Without High School Education",
       x = "% Not High School Graduates",
       y = "Violent Crime Rate")


# Unemployment vs. Crime (95% confidence)
ggplot(data = data_clean, mapping = aes(x = PctUnemployed, y = ViolentCrimesPerPop)) +
  geom_point(alpha = 0.3) +
  geom_smooth(method = "lm", color = "pink") +
  labs(title = "Unemployment vs. Violent Crime",
       x = "% Unemployed", y = "Violent Crime Rate")


# Correlation with Crime 
correlations <- data_clean %>%
  select(-ViolentCrimesPerPop) %>%
  summarise(across(everything(), ~ cor(., data_clean$ViolentCrimesPerPop, use = "complete.obs"))) %>%
  pivot_longer(cols = everything(), names_to = "Variable", values_to = "Correlation") %>%
  arrange(desc(abs(Correlation)))
head(correlations, 30)

correlations %>%
  top_n(5, abs(Correlation)) %>%
  ggplot(aes(x = reorder(Variable, Correlation), y = Correlation, fill = Correlation > 0)) +
  geom_col(show.legend = FALSE) +
  coord_flip() +
  labs(title = "Top 5 Features Most Correlated with Violent Crime",
       x = "Variable", y = "Correlation")

#Income vs. Crime
ggplot(data_clean, aes(x = medIncome, y = ViolentCrimesPerPop)) +
  geom_point(alpha = 0.5, color = "black") +
  geom_smooth(method = "lm", color = "red") +
  labs(title = "Income vs. Violent Crime",
       x = "Median Income",
       y = "Violent Crimes per Population")


```

