---
title: "IDS investigation worksheet"
author: "by 4677: Mate Roka, Xu Sun, Adi Hirasaki, & Xiufeng Zhu"
date: "`r Sys.Date()`"
output: html_document
---

**Note:** You can use this file as you 'working document' where you can try out various investigation ideas and keep notes about your findings. How you use and structure this file is up to you. It is recommended that you keep notes about what you are investigating and what you find as this will make the process of creating your presentation and report easier. Please note that you _do not_ need to submit this file as part of your group project.



```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r load-lib, message = FALSE}
library(tidyverse)
# Add any other libraries here
library(dplyr)
library(tidyr)

```


```{r load-data}


# Load the dataset
data <- read.csv("data/communities.data", stringsAsFactors = FALSE)

# Remove columns with more than 10% missing values
missing <- colMeans(is.na(data)) * 100
data <- data[, missing < 10]

# Drop rows with any remaining missing values
data <- data %>% drop_na()

# Reset row indices
data <- data %>% mutate(row_id = row_number())

# Print the new shape of the cleaned data
cat("Data cleaned. New shape:", nrow(data), "rows and", ncol(data), "columns\n")

# View the cleaned data
head(data)




```


