---
title: "IDS investigation worksheet by Xiufeng"
author: "by 4677:Xiufeng Zhu"
date: "`2025.10.28"
output: html_document
editor_options: 
  markdown: 
    wrap: sentence
---

**Note:** You can use this file as you 'working document' where you can try out various investigation ideas and keep notes about your findings.
How you use and structure this file is up to you.
It is recommended that you keep notes about what you are investigating and what you find as this will make the process of creating your presentation and report easier.
Please note that you *do not* need to submit this file as part of your group project.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r load-lib, message = FALSE}
options(repos = c(CRAN = "https://cloud.r-project.org"))
library(tidyverse)
# Add any other libraries here
library(dplyr)
library(tidyr)


```

Analysis Steps Load the Dataset:

Imported from the CSV file.
Summary statistics reviewed.

Data Cleaning:
Handled missing values.
Removed features with excessive missing data.
Standardized data for regression.

What kind of data does it contain?

The dataset includes over 120 variables, mostly numerical, grouped into several categories:
1. Crime-related variables
Examples:
ViolentCrimesPerPop (main target — violent crimes per 100K population),population
These describe the crime levels in each community.

2. Socioeconomic attributes
Examples:
medIncome (median income),pctWWage
These capture the economic and educational environment.

3. Demographic attributes
Examples:
racePctWhite, racepctBlack, racePctAsian
These describe the population composition of each community.

4. Family characteristics
Examples:
PctIlleg
Reflect family structure and child well-being.

5. Housing characteristics
Examples:

Capture living conditions and housing costs.

6. Law enforcement attributes (LEMAS survey)
Examples:

These describe police department resources and characteristics.

7. Community identifiers

These exist in the original file but are usually removed:


Problem Statement:
The goal of this project is to investigate the factors that are most strongly associated with violent crime rates across U.S. communities using the UCI Communities and Crime dataset. This dataset contains a wide range of socioeconomic, demographic, housing, and policing attributes, all of which may contribute differently to community crime levels.

The primary objectives are:

1.To identify patterns and relationships between community characteristics and violent crime.

2.To determine which variables are the strongest predictors of violent crime rates.

3.To develop and evaluate predictive models (e.g., linear regression and random forest) capable of estimating violent crime rates using community-level attributes.

A secondary objective is to reflect on ethical considerations, such as the potential risks of using demographic variables in crime prediction models.

Exploratory Data Analysis (EDA):
Visualized relationships between features and crime rates.
Identified patterns and key features.

Regression Modeling:
Linear Regression: Modeled continuous crime rates based on features.
Logistic Regression: Converted the crime rate into binary classification (e.g., high/low) for logistic analysis.
Evaluation:

Analyzed model performance metrics (e.g., accuracy, R-squared, confusion matrix).

Classic usage scenarios: In the field of crime analysis and prediction, the classic application scenarios of the Communities and Crime dataset mainly focus on predicting the crime rate of communities through regression models.

Researchers utilize the socio-economic, law enforcement, and demographic data in this dataset to identify the key factors influencing the crime rate and make predictions.
This analysis not only helps understand the drivers of the crime rate but also provides data support for policymakers to formulate more effective prevention strategies.

Solve academic problems: The Communities and Crime dataset addresses a fundamental issue in criminology research: how to quantify and predict the crime rate in communities.
By integrating multi-dimensional socio-economic and demographic data, this dataset provides a rich resource for the academic community to explore the complex relationships between crime rates and various social factors.
This not only drives the research on crime prediction models but also offers scientific basis for decision-making in the fields of social policies and public safety.

Practical application: In practical applications, the Communities and Crime dataset is widely used in urban planning and public safety management.
For instance, local governments and law enforcement agencies can utilize the analysis results of this dataset to optimize resource allocation and enhance community security.
Additionally, non-profit organizations and community groups can also leverage these data to design targeted social intervention programs in order to reduce crime rates and improve the community environment.



| Type | What to Add                                | Why It Improves Your Project                              |
| ---- | ------------------------------------------ | --------------------------------------------------------- |
| 1️⃣  | Correlation heatmap                        | Visualize variable relationships clearly                  |
| 2️⃣  | Distribution plots                         | Show how target and predictors vary                       |
| 3️⃣  | Variable selection (feature reduction)     | Simplify model and improve interpretability               |
| 4️⃣  | Cross-validation                           | Make model evaluation more robust                         |
| 5️⃣  | Residual analysis                          | Check model errors for bias                               |
| 6️⃣  | Map or geographic visualization (optional) | Show crime rates per area (if you find state/county info) |
| 7️⃣  | Model comparison plot                      | Visually compare models                                   |
| 8️⃣  | Deeper ethics discussion                   | Shows awareness of real-world implications                |

> **Ethical Considerations**  
> The Communities and Crime dataset contains sensitive attributes such as race, income, and family structure.  
> These variables can reflect systemic inequalities and should be treated carefully:
> - Avoid using models like these for **individual prediction or policing**.  
> - Be aware of **bias amplification**: if biased data (e.g., due to unequal policing) is used to train a model, predictions will also be biased.  
> - Focus on **understanding community-level patterns** and informing equitable policy decisions rather than punitive actions.  
>  
> Ethical data science means asking not just *“Can we predict it?”* but also *“Should we?”*.



```{r load-data}
library(tidyverse)
options(repos = c(CRAN = "https://cloud.r-project.org"))

# Load column names
names_url <- "https://archive.ics.uci.edu/ml/machine-learning-databases/communities/communities.names"
name_lines <- readLines(names_url)
name_lines <- name_lines[grepl("^@attribute", name_lines)]
col_names <- gsub("^@attribute\\s+", "", name_lines)
col_names <- sub("\\s+.*$", "", col_names)
col_names <- col_names[col_names != ""]

# Load dataset
data_url <- "https://archive.ics.uci.edu/ml/machine-learning-databases/communities/communities.data"
crime <- read.csv(data_url, header = FALSE, na.strings = "?", col.names = col_names)

# Clean dataset
crime <- crime %>%
  select(-state, -county, -community, -communityname, -fold)

# Replace missing values with column means
library(dplyr)

data_clean <- crime %>%
  mutate(across(where(is.numeric),
                ~ ifelse(is.na(.), mean(., na.rm = TRUE), .)))
    # why? -adi

# Print the new shape of the cleaned data
summary(data_clean)
cat("Data cleaned. Shape:", nrow(data_clean), "rows and", ncol(data_clean), "columns\n")

library(corrplot)

# Compute correlations among numeric columns
crime_num <- data_clean %>% select(where(is.numeric))
corr_matrix <- cor(crime_num, use = "complete.obs")

# Plot heatmap for top correlated features
corrplot(corr_matrix, type = "lower", tl.cex = 0.2, tl.col = "black")

# Target Variable Distribution
ggplot(data_clean, aes(x = ViolentCrimesPerPop)) +
  geom_histogram(bins = 30, fill = "steelblue", color = "white") +
  labs(title = "Distribution of Violent Crime Rate",
       x = "Violent Crimes per Population", y = "Frequency")

ggplot(data_clean, aes(x = PctPopUnderPov, y = ViolentCrimesPerPop)) +
geom_point(alpha = 0.5, color = "purple") +
geom_smooth(method = "lm", se = FALSE, color = "red") +
labs(title = "Violent Crime vs Poverty", x = "Poverty (%)", y = "Violent Crime Rate")

library(caret)      
library(randomForest) 

set.seed(123)
train_index <- createDataPartition(data_clean$ViolentCrimesPerPop, p = 0.8, list = FALSE)
train <- data_clean[train_index, ]
test  <- data_clean[-train_index, ]

# Correlation with target
corrs <- sapply(data_clean, function(x) cor(x, data_clean$ViolentCrimesPerPop, use = "complete.obs"))
head(sort(corrs, decreasing = TRUE), 10)

summary(data_clean$ViolentCrimesPerPop)

# Check a few relationships
plot(data_clean$medIncome, data_clean$ViolentCrimesPerPop,
     main = "Crime vs Median Income",
     xlab = "Median Income", ylab = "Violent Crime Rate")

plot(data_clean$pctWWage, data_clean$ViolentCrimesPerPop,
     main = "Crime vs WWage",
     xlab = "WWage %", ylab = "Violent Crime Rate")

# Build two models

# (a) Linear Regression
lm_model <- lm(ViolentCrimesPerPop ~ ., data = train)
lm_pred <- predict(lm_model, newdata = test)

# (b) Random Forest
rf_model <- randomForest(ViolentCrimesPerPop ~ ., data = train, ntree = 150)
rf_pred <- predict(rf_model, newdata = test)

# Evaluate models
# Calculate R-squared (how well model fits) and RMSE (error)
lm_r2 <- cor(lm_pred, test$ViolentCrimesPerPop)^2
rf_r2 <- cor(rf_pred, test$ViolentCrimesPerPop)^2

lm_rmse <- sqrt(mean((lm_pred - test$ViolentCrimesPerPop)^2))
rf_rmse <- sqrt(mean((rf_pred - test$ViolentCrimesPerPop)^2))

cat("Linear Regression: R2 =", lm_r2, " RMSE =", lm_rmse, "\n")
cat("Random Forest: R2 =", rf_r2, " RMSE =", rf_rmse, "\n")


# Visualize predictions 
results <- data.frame(
  Actual = test$ViolentCrimesPerPop,
  Predicted = rf_pred
)

ggplot(results, aes(x = Actual, y = Predicted)) +
  geom_point(color = "blue", alpha = 0.5) +
  geom_abline(slope = 1, intercept = 0, color = "red", linetype = "dashed") +
  labs(title = "Random Forest Predictions vs Actual",
       x = "Actual Violent Crime Rate",
       y = "Predicted Crime Rate")

# Feature importance (which factors matter most) 
varImpPlot(rf_model, n.var = 10, main = "Most Important Features")

# Residual Analysis (Model Diagnostics)
residuals_rf <- test$ViolentCrimesPerPop - rf_pred

ggplot(data.frame(residuals_rf), aes(x = residuals_rf)) +
  geom_histogram(bins = 30, fill = "darkorange", color = "white") +
  labs(title = "Residual Distribution (Random Forest)",
       x = "Residuals", y = "Count")

library(plotly)
p <- ggplot(crime, aes(x = PctPopUnderPov, y = ViolentCrimesPerPop)) +
  geom_point(alpha = 0.6, color = "purple") +
  labs(title = "Interactive: Poverty vs Violent Crime",
       x = "Poverty (%)", y = "Violent Crime Rate")


ggplotly(p)




```


```{r}
# --- Community Crime Analysis in R ---
#
# This script replicates the Python analysis using the 'tidymodels' framework.
#
# Before running, you must install the required packages:
# install.packages(c("readr", "dplyr", "ggplot2", "tidymodels", "ranger", "caTools"))
#
# --- 1. Load Libraries ---

# This list of column names is from the dataset's documentation
# NOTE: Corrected a typo from the original Python list ("Good," was removed)
column_names <- c(
    'state', 'county', 'community', 'communityname', 'fold', 'population', 'householdsize',
    'racepctblack', 'racePctWhite', 'racePctAsian', 'racePctHisp', 'agePct12t21',
    'agePct12t29', 'agePct16t24', 'agePct65up', 'numbUrban', 'pctUrban', 'medIncome',
    'pctWWage', 'pctWFarmSelf', 'pctWInvInc', 'pctWSocSec', 'pctWPubAsst', 'pctWRetire',
    'medFamInc', 'perCapInc', 'whitePerCap', 'blackPerCap', 'indianPerCap', 'AsianPerCap',
    'OtherPerCap', 'HispPerCap', 'NumUnderPov', 'PctNotHSGrad', 'PctBSorMore',
    'PctUnemployed', 'PctEmploy', 'PctEmplManu', 'PctEmplProfServ', 'PctOccupManu',
    'PctOccupProfMgmt', 'MalePctDivorce', 'MalePctNevMarr', 'FemalePctDiv', 'TotalPctDiv',
    'PersPerFam', 'PctFam2Par', 'PctKids2Par', 'PctYoungKids2Par', 'PctTeen2Par',
    'PctWorkMomYoungKids', 'PctWorkMom', 'NumIlleg', 'PctIlleg', 'NumImmig',
    'PctImmigRecent', 'PctImmigRec5', 'PctImmigRec8', 'PctImmigRec10', 'PctRecentImmig',
    'PctRecImmig5', 'PctRecImmig8', 'PctRecImmig10', 'PctSpeakEnglOnly',
    'PctNotSpeakEnglWell', 'PctLargHouseFam', 'PctLargHouseOccup', 'PersPerOccupHous',
    'PersPerOwnOccHous', 'PersPerRentOccHous', 'PctPersOwnOccup', 'PctPersDenseHous',
    'PctHousLess3BR', 'MedNumBR', 'HousVacant', 'PctHousOccup', 'PctHousOwnOcc',
    'PctVacantBoarded', 'PctVacMore6Mos', 'MedYrHousBuilt', 'PctHousNoPhone',
    'PctWOFullPlumb', 'OwnOccLowQuart', 'OwnOccMedVal', 'OwnOccHiQuart', 'RentLowQ',
    'RentMedian', 'RentHighQ', 'MedRent', 'MedRentPctHousInc', 'MedOwnCostPctInc',
    'MedOwnCostPctIncNoMtg', 'NumInShelters', 'NumStreet', 'PctForeignBorn',
    'PctBornSameState', 'PctSameHouse85', 'PctSameCity85', 'PctSameState85',
    'LemasSwornFT', 'LemasSwFTPerPop', 'LemasSwFTFieldOps', 'LemasSwFTFieldPerPop',
    'LemasTotalReq', 'LemasTotReqPerPop', 'PolicReqPerPop', 'PolicPerPop',
    'RacialMatchCommPol', 'PctPolicWhite', 'PctPolicBlack', 'PctPolicHisp',
    'PctPolicAsian', 'PctPolicMinor', 'OfficAssgnDrugUnits', 'NumKindsDrugsSeiz',
    'PolicPerPopAgg', 'PolicFieldPerPopAgg', 'LemasPctPolicOnPatr',
    'LemasGangUnitDeploy', 'LemasPctOfficDrugUn', 'PolicBudgPerPop',
    'ViolentCrimesPerPop' # This is the target variable
)

load_data <- function(filename = 'communities.data') {
  # Loads the communities.data file, replacing '?' with NA.
  cat(sprintf("Loading data from %s...\n", filename))
  
  # The dataset uses '?' for missing values
  data <- read_csv(filename, col_names = column_names, na = '?', col_types = cols())
  
  cat(sprintf("Data loaded: %d rows, %d columns\n", nrow(data), ncol(data)))
  return(data)
}

preprocess_and_split_data <- function(data) {
  # Prepares the data for modeling:
  # 1. Drops non-predictive columns.
  # 2. Drops rows where the target variable is missing.
  # 3. Drops feature columns with more than 50% missing values.
  # 4. Splits into training and test sets.
  
  cat("Preprocessing data...\n")
  
  # 1. Drop non-predictive columns
  non_predictive_cols <- c('state', 'county', 'community', 'communityname', 'fold')
  data <- data %>% select(-all_of(non_predictive_cols))
  
  # 2. Drop rows with missing target variable
  data <- data %>% filter(!is.na(ViolentCrimesPerPop))
  
  # 3. Drop feature columns with >50% missing values
  # Calculate missing percentage for each column
  missing_pct <- sapply(data, function(x) mean(is.na(x)))
  
  # Identify columns to drop (excluding the target)
  cols_to_drop <- names(missing_pct)[missing_pct > 0.5 & names(missing_pct) != "ViolentCrimesPerPop"]
  
  data <- data %>% select(-all_of(cols_to_drop))
  cat(sprintf("Dropped %d columns due to >50%% missing values: %s\n", 
              length(cols_to_drop), paste(cols_to_drop, collapse=", ")))
              
  # 4. Split into training and test sets
  set.seed(42) # For reproducibility
  split <- sample.split(data$ViolentCrimesPerPop, SplitRatio = 0.8)
  train_data <- subset(data, split == TRUE)
  test_data <- subset(data, split == FALSE)
  
  cat(sprintf("Training set: %d samples\n", nrow(train_data)))
  cat(sprintf("Test set: %d samples\n", nrow(test_data)))
  
  return(list(train = train_data, test = test_data))
}

perform_eda <- function(train_data) {
  # Performs the EDA task: finding top correlated features.
  cat("\n--- Exploratory Data Analysis (EDA) ---\n")
  
  # Use 'pairwise.complete.obs' to handle missing values during correlation calculation
  cor_matrix <- cor(train_data, use = "pairwise.complete.obs")
  
  # Get correlation with target, remove NA, sort by absolute value
  cor_with_target <- cor_matrix[, "ViolentCrimesPerPop"]
  cor_with_target <- cor_with_target[!is.na(cor_with_target)]
  cor_with_target_abs_sorted <- sort(abs(cor_with_target), decreasing = TRUE)
  
  cat("Top 15 Features Most Correlated with Violent Crime:\n")
  print(cor_with_target_abs_sorted[2:16]) # [1] is the target itself
}

build_and_evaluate_models <- function(train_data, test_data) {
  # Builds, trains, and evaluates models using a tidymodels workflow.
  cat("\n--- Model Building & Evaluation ---\n")

  # 1. Define the preprocessing recipe
  # This is the R equivalent of the sklearn preprocessor pipeline
  data_recipe <- recipe(ViolentCrimesPerPop ~ ., data = train_data) %>%
    step_impute_median(all_numeric_predictors()) %>% # Fill missing with median
    step_scale(all_numeric_predictors()) # Standardize (center and scale)
    # We add step_dummy(all_nominal_predictors()) if we had text/factor columns

  # 2. Define the model specifications
  # Linear Regression
  lr_spec <- linear_reg() %>%
    set_engine("lm")
  
  # Random Forest (using 'ranger' engine)
  rf_spec <- rand_forest(trees = 100) %>%
    set_engine("ranger", importance = "permutation") %>% # 'permutation' is a robust way to get feature importance
    set_mode("regression")

  # 3. Create the workflows (Pipeline = Recipe + Model)
  lr_workflow <- workflow() %>%
    add_recipe(data_recipe) %>%
    add_model(lr_spec)
  
  rf_workflow <- workflow() %>%
    add_recipe(data_recipe) %>%
    add_model(rf_spec)

  # 4. Train the models
  cat("Training Linear Regression model...\n")
  lr_fit <- fit(lr_workflow, data = train_data)
  
  cat("Training Random Forest model...\n")
  rf_fit <- fit(rf_workflow, data = train_data)

  # 5. Evaluate the models on the test set
  cat("\n--- Model Results ---\n")
  
  # Define the metrics we care about
  model_metrics <- metric_set(rsq, rmse)
  
  # Evaluate Linear Regression
  lr_preds <- predict(lr_fit, new_data = test_data) %>%
    bind_cols(test_data %>% select(ViolentCrimesPerPop))
  
  lr_results <- lr_preds %>%
    model_metrics(truth = ViolentCrimesPerPop, estimate = .pred)
  
  cat("Linear Regression:\n")
  print(lr_results)
  
  # Evaluate Random Forest
  rf_preds <- predict(rf_fit, new_data = test_data) %>%
    bind_cols(test_data %>% select(ViolentCrimesPerPop))
  
  rf_results <- rf_preds %>%
    model_metrics(truth = ViolentCrimesPerPop, estimate = .pred)
  
  cat("\nRandom Forest Regression:\n")
  print(rf_results)
  
  # Return the fitted RF model and both sets of predictions
  return(list(fit_rf = rf_fit, predictions_rf = rf_preds, predictions_lr = lr_preds))
}

get_feature_importance <- function(rf_fit_workflow) {
  # Extracts and prints feature importances from the fitted Random Forest.
  cat("\n--- Random Forest Feature Importance ---\n")
  
  # Extract the fitted 'ranger' object from the workflow
  rf_model <- extract_fit_parsnip(rf_fit_workflow)$fit
  
  # Get importance, convert to data frame, sort, and print top 10
  importances <- data.frame(
    feature = names(rf_model$variable.importance),
    importance = rf_model$variable.importance
  ) %>%
    arrange(desc(importance))
    
  cat("Top 10 Most Important Features:\n")
  print(head(importances, 10))
}

visualize_results <- function(rf_predictions, lr_predictions) {
  # Generates the visualizations described in the report using ggplot2.
  cat("\n--- Generating Visualizations ---\n")
  
  # --- Linear Regression Plots ---
  
  # 1. Predicted vs. Actual Plot (Linear Regression)
  plot1_lr <- ggplot(lr_predictions, aes(x = ViolentCrimesPerPop, y = .pred)) +
    # MODIFIED: Increased size to 2.0 and set alpha to 1.0 (fully opaque)
    geom_point(alpha = 1.0, size = 2.0, color = "blue") +
    geom_abline(slope = 1, intercept = 0, color = "red", linetype = "dashed", size = 1) +
    labs(
      title = "Linear Regression: Predicted vs. Actual Values (Test Set)",
      x = "Actual Violent Crimes Per Pop",
      y = "Predicted Violent Crimes Per Pop"
    ) +
    theme_minimal() +
    coord_fixed()
  
  ggsave("predicted_vs_actual_LR.png", plot1_lr)
  cat("Saved 'predicted_vs_actual_LR.png'\n")

  # 2. Residual Distribution Map (Linear Regression)
  lr_predictions <- lr_predictions %>%
    mutate(residuals = ViolentCrimesPerPop - .pred)
    
  plot2_lr <- ggplot(lr_predictions, aes(x = residuals)) +
    # Histogram uses same fill color as previous version
    geom_histogram(bins = 30, fill = "steelblue", color = "black", alpha = 0.7) +
    geom_vline(xintercept = 0, color = "red", linetype = "dashed", size = 1) +
    labs(
      title = "Linear Regression: Histogram of Residuals",
      x = "Residual (Actual - Predicted)",
      y = "Frequency"
    ) +
    theme_minimal()
    
  ggsave("residuals_histogram_LR.png", plot2_lr)
  cat("Saved 'residuals_histogram_LR.png'\n")

  # --- Random Forest Plots ---

  # 1. Predicted vs. Actual Plot (Random Forest)
  plot1_rf <- ggplot(rf_predictions, aes(x = ViolentCrimesPerPop, y = .pred)) +
    # MODIFIED: Increased size to 2.0 and set alpha to 1.0 (fully opaque)
    geom_point(alpha = 1.0, size = 2.0, color = "darkgreen") +
    geom_abline(slope = 1, intercept = 0, color = "red", linetype = "dashed", size = 1) +
    labs(
      title = "Random Forest: Predicted vs. Actual Values (Test Set)",
      x = "Actual Violent Crimes Per Pop",
      y = "Predicted Violent Crimes Per Pop"
    ) +
    theme_minimal() +
    coord_fixed() # Ensures a 1:1 aspect ratio
  
  ggsave("predicted_vs_actual_R.png", plot1_rf)
  cat("Saved 'predicted_vs_actual_R.png'\n")

  # 2. Residual Distribution Map (Random Forest)
  # First, calculate residuals
  rf_predictions <- rf_predictions %>%
    mutate(residuals = ViolentCrimesPerPop - .pred)
    
  plot2_rf <- ggplot(rf_predictions, aes(x = residuals)) +
    # Histogram uses same fill color as previous version
    geom_histogram(bins = 30, fill = "darkgreen", color = "black", alpha = 0.7) +
    geom_vline(xintercept = 0, color = "red", linetype = "dashed", size = 1) +
    labs(
      title = "Random Forest: Histogram of Residuals",
      x = "Residual (Actual - Predicted)",
      y = "Frequency"
    ) +
    theme_minimal()
    
  ggsave("residuals_histogram_R.png", plot2_rf)
  cat("Saved 'residuals_histogram_R.png'\n")
}

# --- Main execution ---
main <- function() {
  # Ensure the data file 'communities.data' is in the same directory
  data <- tryCatch({
    load_data('communities.data')
  }, error = function(e) {
    cat("Error: 'communities.data' file not found.\n")
    cat("Please download it from the UCI repository or upload it.\n")
    return(NULL)
  })
  
  if (!is.null(data)) {
    data_splits <- preprocess_and_split_data(data)
    
    perform_eda(data_splits$train)
    
    # Capture results for both models
    model_results <- build_and_evaluate_models(data_splits$train, data_splits$test)
    
    get_feature_importance(model_results$fit_rf)
    
    # Pass predictions for both models to visualization
    visualize_results(model_results$predictions_rf, model_results$predictions_lr)
    
    cat("\nAnalysis complete.\n")
  }
}

# Run the main function
main()
```


```{r}
# --- Community Crime Analysis (Simplified R Script) ---
#
# This script performs the entire analysis (Data Prep, EDA, Model Building, Evaluation, and Visualization)
# without custom functions, running sequentially for maximum readability.
#
# Prerequisite: Ensure you have the 'communities.data' file in the same directory.
#
# Before running, you must install the required packages:
# install.packages(c("readr", "dplyr", "ggplot2", "tidymodels", "ranger", "caTools"))
#
# ----------------------------------------------------------------------

# --- 1. Load Libraries ---
library(readr)
library(dplyr)
library(ggplot2)
library(tidymodels)
library(ranger)
library(caTools) # For sample.split

cat("Libraries loaded successfully.\n")

# --- 2. Define Column Names ---
# This list is necessary as the data file does not contain a header row.
column_names <- c(
    'state', 'county', 'community', 'communityname', 'fold', 'population', 'householdsize',
    'racepctblack', 'racePctWhite', 'racePctAsian', 'racePctHisp', 'agePct12t21',
    'agePct12t29', 'agePct16t24', 'agePct65up', 'numbUrban', 'pctUrban', 'medIncome',
    'pctWWage', 'pctWFarmSelf', 'pctWInvInc', 'pctWSocSec', 'pctWPubAsst', 'pctWRetire',
    'medFamInc', 'perCapInc', 'whitePerCap', 'blackPerCap', 'indianPerCap', 'AsianPerCap',
    'OtherPerCap', 'HispPerCap', 'NumUnderPov', 'PctNotHSGrad', 'PctBSorMore',
    'PctUnemployed', 'PctEmploy', 'PctEmplManu', 'PctEmplProfServ', 'PctOccupManu',
    'PctOccupProfMgmt', 'MalePctDivorce', 'MalePctNevMarr', 'FemalePctDiv', 'TotalPctDiv',
    'PersPerFam', 'PctFam2Par', 'PctKids2Par', 'PctYoungKids2Par', 'PctTeen2Par',
    'PctWorkMomYoungKids', 'PctWorkMom', 'NumIlleg', 'PctIlleg', 'NumImmig',
    'PctImmigRecent', 'PctImmigRec5', 'PctImmigRec8', 'PctImmigRec10', 'PctRecentImmig',
    'PctRecImmig5', 'PctRecImmig8', 'PctRecImmig10', 'PctSpeakEnglOnly',
    'PctNotSpeakEnglWell', 'PctLargHouseFam', 'PctLargHouseOccup', 'PersPerOccupHous',
    'PersPerOwnOccHous', 'PersPerRentOccHous', 'PctPersOwnOccup', 'PctPersDenseHous',
    'PctHousLess3BR', 'MedNumBR', 'HousVacant', 'PctHousOccup', 'PctHousOwnOcc',
    'PctVacantBoarded', 'PctVacMore6Mos', 'MedYrHousBuilt', 'PctHousNoPhone',
    'PctWOFullPlumb', 'OwnOccLowQuart', 'OwnOccMedVal', 'OwnOccHiQuart', 'RentLowQ',
    'RentMedian', 'RentHighQ', 'MedRent', 'MedRentPctHousInc', 'MedOwnCostPctInc',
    'MedOwnCostPctIncNoMtg', 'NumInShelters', 'NumStreet', 'PctForeignBorn',
    'PctBornSameState', 'PctSameHouse85', 'PctSameCity85', 'PctSameState85',
    'LemasSwornFT', 'LemasSwFTPerPop', 'LemasSwFTFieldOps', 'LemasSwFTFieldPerPop',
    'LemasTotalReq', 'LemasTotReqPerPop', 'PolicReqPerPop', 'PolicPerPop',
    'RacialMatchCommPol', 'PctPolicWhite', 'PctPolicBlack', 'PctPolicHisp',
    'PctPolicAsian', 'PctPolicMinor', 'OfficAssgnDrugUnits', 'NumKindsDrugsSeiz',
    'PolicPerPopAgg', 'PolicFieldPerPopAgg', 'LemasPctPolicOnPatr',
    'LemasGangUnitDeploy', 'LemasPctOfficDrugUn', 'PolicBudgPerPop',
    'ViolentCrimesPerPop' # The target variable
)

# --- 3. Load Data ---
filename <- 'communities.data'
cat(sprintf("Loading data from %s...\n", filename))

# The dataset uses '?' for missing values, so we specify this in read_csv
data_raw <- read_csv(filename, col_names = column_names, na = '?', col_types = cols())

cat(sprintf("Data loaded: %d rows, %d columns\n", nrow(data_raw), ncol(data_raw)))

# --- 4. Data Preprocessing (Cleaning) ---
cat("\nPreprocessing data (dropping unnecessary and incomplete features)...\n")

# Drop non-predictive columns (metadata)
non_predictive_cols <- c('state', 'county', 'community', 'communityname', 'fold')
data_clean <- data_raw %>% select(-all_of(non_predictive_cols))

# Drop rows where the target variable is missing (not useful for training)
data_clean <- data_clean %>% filter(!is.na(ViolentCrimesPerPop))

# Drop feature columns with >50% missing values
missing_pct <- sapply(data_clean, function(x) mean(is.na(x)))
cols_to_drop <- names(missing_pct)[missing_pct > 0.5 & names(missing_pct) != "ViolentCrimesPerPop"]
data_clean <- data_clean %>% select(-all_of(cols_to_drop))

cat(sprintf("Total columns after initial cleaning: %d\n", ncol(data_clean)))

# --- 5. Split Data into Training and Testing Sets ---
set.seed(42) # For reproducibility
split <- sample.split(data_clean$ViolentCrimesPerPop, SplitRatio = 0.8)
train_data <- subset(data_clean, split == TRUE)
test_data <- subset(data_clean, split == FALSE)

cat(sprintf("Training set: %d samples\n", nrow(train_data)))
cat(sprintf("Test set: %d samples\n", nrow(test_data)))

# --- 6. Exploratory Data Analysis (EDA) - Correlation ---
cat("\n--- Exploratory Data Analysis (EDA) ---\n")

# Calculate correlation matrix
cor_matrix <- cor(train_data, use = "pairwise.complete.obs")

# Get correlation with target variable, sort by absolute value
cor_with_target <- cor_matrix[, "ViolentCrimesPerPop"]
cor_with_target <- cor_with_target[!is.na(cor_with_target)]
cor_with_target_abs_sorted <- sort(abs(cor_with_target), decreasing = TRUE)

cat("Top 15 Features Most Correlated with Violent Crime:\n")
# Print features 2 through 16 (1 is the target itself)
print(cor_with_target_abs_sorted[2:16])

# --- 7. Define Preprocessing Recipe (Imputation and Scaling) ---
# The 'recipe' handles the final missing value imputation (median) and standardization (scaling)
data_recipe <- recipe(ViolentCrimesPerPop ~ ., data = train_data) %>%
  step_impute_median(all_numeric_predictors()) %>% # Fill any remaining NA with median
  step_scale(all_numeric_predictors())           # Standardize features

# --- 8. Define Model Specifications ---

# Linear Regression (Simple Benchmark Model)
lr_spec <- linear_reg() %>%
  set_engine("lm")

# Random Forest (Complex Ensemble Model)
rf_spec <- rand_forest(trees = 100) %>%
  set_engine("ranger", importance = "permutation") %>%
  set_mode("regression")

# --- 9. Create and Train Workflows ---

# Linear Regression Workflow
lr_workflow <- workflow() %>%
  add_recipe(data_recipe) %>%
  add_model(lr_spec)

cat("\nTraining Linear Regression model...\n")
lr_fit <- fit(lr_workflow, data = train_data)

# Random Forest Workflow
rf_workflow <- workflow() %>%
  add_recipe(data_recipe) %>%
  add_model(rf_spec)

cat("Training Random Forest model...\n")
rf_fit <- fit(rf_workflow, data = train_data)

# --- 10. Evaluate Models on Test Set ---
cat("\n--- Model Results ---\n")

# Define the metrics (R-squared and RMSE)
model_metrics <- metric_set(rsq, rmse)

# Get Predictions for Linear Regression
lr_preds <- predict(lr_fit, new_data = test_data) %>%
  bind_cols(test_data %>% select(ViolentCrimesPerPop))

# Calculate and print LR metrics
lr_results <- lr_preds %>%
  model_metrics(truth = ViolentCrimesPerPop, estimate = .pred)

cat("Linear Regression Performance:\n")
print(lr_results)

# Get Predictions for Random Forest
rf_preds <- predict(rf_fit, new_data = test_data) %>%
  bind_cols(test_data %>% select(ViolentCrimesPerPop))

# Calculate and print RF metrics
rf_results <- rf_preds %>%
  model_metrics(truth = ViolentCrimesPerPop, estimate = .pred)

cat("\nRandom Forest Regression Performance:\n")
print(rf_results)

# --- 11. Random Forest Feature Importance ---
cat("\n--- Random Forest Feature Importance ---\n")

# Extract the fitted model and importance scores
rf_model <- extract_fit_parsnip(rf_fit)$fit
importances <- data.frame(
  feature = names(rf_model$variable.importance),
  importance = rf_model$variable.importance
) %>%
  arrange(desc(importance))

cat("Top 10 Most Important Features:\n")
print(head(importances, 10))

# --- 12. Visualization (Predicted vs. Actual and Residuals) ---
cat("\n--- Generating Visualizations ---\n")

# Calculate residuals for both models
lr_preds <- lr_preds %>% mutate(residuals = ViolentCrimesPerPop - .pred)
rf_preds <- rf_preds %>% mutate(residuals = ViolentCrimesPerPop - .pred)

# --- 12.1 Predicted vs. Actual Plot (Linear Regression) ---
plot1_lr <- ggplot(lr_preds, aes(x = ViolentCrimesPerPop, y = .pred)) +
  geom_point(alpha = 1.0, size = 2.0, color = "blue") +
  geom_abline(slope = 1, intercept = 0, color = "red", linetype = "dashed", size = 1) +
  labs(
    title = "Linear Regression: Predicted vs. Actual Values",
    x = "Actual Crime Rate",
    y = "Predicted Crime Rate"
  ) +
  theme_minimal() +
  coord_fixed()
  
ggsave("predicted_vs_actual_LR.png", plot1_lr, width = 6, height = 6)
cat("Saved 'predicted_vs_actual_LR.png' \n")

# --- 12.2 Residual Distribution Map (Linear Regression) ---
plot2_lr <- ggplot(lr_preds, aes(x = residuals)) +
  geom_histogram(bins = 30, fill = "steelblue", color = "black", alpha = 0.7) +
  geom_vline(xintercept = 0, color = "red", linetype = "dashed", size = 1) +
  labs(
    title = "Linear Regression: Residual Distribution",
    x = "Residual (Error)",
    y = "Frequency"
  ) +
  theme_minimal()
  
ggsave("residuals_histogram_LR.png", plot2_lr, width = 6, height = 4)
cat("Saved 'residuals_histogram_LR.png' \n")

# --- 12.3 Predicted vs. Actual Plot (Random Forest) ---
plot1_rf <- ggplot(rf_preds, aes(x = ViolentCrimesPerPop, y = .pred)) +
  geom_point(alpha = 1.0, size = 2.0, color = "darkgreen") +
  geom_abline(slope = 1, intercept = 0, color = "red", linetype = "dashed", size = 1) +
  labs(
    title = "Random Forest: Predicted vs. Actual Values",
    x = "Actual Crime Rate",
    y = "Predicted Crime Rate"
  ) +
  theme_minimal() +
  coord_fixed()
  
ggsave("predicted_vs_actual_R.png", plot1_rf, width = 6, height = 6)
cat("Saved 'predicted_vs_actual_R.png' \n")

# --- 12.4 Residual Distribution Map (Random Forest) ---
plot2_rf <- ggplot(rf_preds, aes(x = residuals)) +
  geom_histogram(bins = 30, fill = "darkgreen", color = "black", alpha = 0.7) +
  geom_vline(xintercept = 0, color = "red", linetype = "dashed", size = 1) +
  labs(
    title = "Random Forest: Residual Distribution (Sharply Centered at Zero)",
    x = "Residual (Error)",
    y = "Frequency"
  ) +
  theme_minimal()
  
ggsave("residuals_histogram_R.png", plot2_rf, width = 6, height = 4)
cat("Saved 'residuals_histogram_R.png' \n")

cat("\nAnalysis complete. Check the output directory for four saved plot files.\n")
# ----------------------------------------------------------------------
```

