---
title: "IDS investigation worksheet"
author: "by 4677: Mate Roka, Xu Sun, Adi Hirasaki, & Xiufeng Zhu"
date: "`r Sys.Date()`"
output: html_document
---

**Note:** You can use this file as you 'working document' where you can try out various investigation ideas and keep notes about your findings. How you use and structure this file is up to you. It is recommended that you keep notes about what you are investigating and what you find as this will make the process of creating your presentation and report easier. Please note that you _do not_ need to submit this file as part of your group project.



```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r load-lib, message = FALSE}
library(tidyverse)
# Add any other libraries here
library(dplyr)
library(tidyr)

```
Analysis Steps
Load the Dataset:

Imported from the CSV file.
Summary statistics reviewed.
Data Cleaning:

Handled missing values.
Removed features with excessive missing data.
Standardized data for regression.
Exploratory Data Analysis (EDA):

Visualized relationships between features and crime rates.
Identified patterns and key features.
Regression Modeling:

Linear Regression:
Modeled continuous crime rates based on features.
Logistic Regression:
Converted the crime rate into binary classification (e.g., high/low) for logistic analysis.
Evaluation:

Analyzed model performance metrics (e.g., accuracy, R-squared, confusion matrix).

Classic usage scenarios:
In the field of crime analysis and prediction, the classic application scenarios of the Communities and Crime dataset mainly focus on predicting the crime rate of communities through regression models. Researchers utilize the socio-economic, law enforcement, and demographic data in this dataset to identify the key factors influencing the crime rate and make predictions. This analysis not only helps understand the drivers of the crime rate but also provides data support for policymakers to formulate more effective prevention strategies.

Solve academic problems:
The Communities and Crime dataset addresses a fundamental issue in criminology research: how to quantify and predict the crime rate in communities. By integrating multi-dimensional socio-economic and demographic data, this dataset provides a rich resource for the academic community to explore the complex relationships between crime rates and various social factors. This not only drives the research on crime prediction models but also offers scientific basis for decision-making in the fields of social policies and public safety.

Practical application:
In practical applications, the Communities and Crime dataset is widely used in urban planning and public safety management. For instance, local governments and law enforcement agencies can utilize the analysis results of this dataset to optimize resource allocation and enhance community security. Additionally, non-profit organizations and community groups can also leverage these data to design targeted social intervention programs in order to reduce crime rates and improve the community environment.
```{r load-data}


# Load the dataset
data <- read.csv("data/communities.data", stringsAsFactors = FALSE)

# Remove columns with more than 10% missing values
missing <- colMeans(is.na(data)) * 100
data <- data[, missing < 10]

# Drop rows with any remaining missing values
data <- data %>% drop_na()

# Reset row indices
data <- data %>% mutate(row_id = row_number())

# Print the new shape of the cleaned data
cat("Data cleaned. New shape:", nrow(data), "rows and", ncol(data), "columns\n")

# View the cleaned data
head(data)






```


