---
title: "investigation-adi"
author: "Adi"
date: "2025-10-17"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = "/Users/Addie/Intro to Data Science/FinalProject")
library(tidyverse)
library(dplyr)
library(tidyr)
```

When you click the **Knit** button a document will be generated that includes both content as well as the output of any embedded R code chunks within the document. You can embed an R code chunk like this:

```{r dataframe setup}
communities <- read_csv("data/communities.data")
communities <- setNames(communities, c("State","County","Community","CommunityName","Fold","Population","HouseholdSize","RacePctBlack","RacePctWhite","RacePctAsian","RacePctHisp","AgePct12t21","AgePct12t29","AgePct16t24","AgePct65up","NumbUrban","PctUrban","MedIncome","PctWWage","PctWFarmSelf","PctWInvInc","PctWSocSec","PctWPubAsst","PctWRetire","MedFamInc","PerCapInc","WhitePerCap","BlackPerCap","IndianPerCap","AsianPerCap","OtherPerCap","HispPerCap","NumUnderPov","PctPopUnderPov","PctLess9thGrade","PctNotHSGrad","PctBSorMore","PctUnemployed","PctEmploy","PctEmplManu","PctEmplProfServ","PctOccupManu","PctOccupMgmtProf","MalePctDivorce","MalePctNevMarr","FemalePctDiv","TotalPctDiv","PersPerFam","PctFam2Par","PctKids2Par","PctYoungKids2Par","PctTeen2Par","PctWorkMomYoungKids","PctWorkMom","NumIlleg","PctIlleg","NumImmig","PctImmigRecent","PctImmigRec5","PctImmigRec8","PctImmigRec10","PctRecentImmig","PctRecImmig5","PctRecImmig8","PctRecImmig10","PctSpeakEnglOnly","PctNotSpeakEnglWell","PctLargHouseFam","PctLargHouseOccup","PersPerOccupHous","PersPerOwnOccHous","PersPerRentOccHous","PctPersOwnOccup","PctPersDenseHous","PctHousLess3BR","MedNumBR","HousVacant","PctHousOccup","PctHousOwnOcc","PctVacantBoarded","PctVacMore6Mos","MedYrHousBuilt","PctHousNoPhone","PctWOFullPlumb","OwnOccLowQuart","OwnOccMedVal","OwnOccHiQuart","RentLowQ","RentMedian","RentHighQ","MedRent","MedRentPctHousInc","MedOwnCostPctInc","MedOwnCostPctIncNoMtg","NumInShelters","NumStreet","PctForeignBorn","PctBornSameState","PctSameHouse85","PctSameCity85","PctSameState85","LemasSwornFT","LemasSwFTPerPop","LemasSwFTFieldOps","LemasSwFTFieldPerPop","LemasTotalReq","LemasTotReqPerPop","PolicRepPerOffic","PolicPerPop","RacialMatchCommPol","PctPolicWhite","PctPolicBlack","PctPolicHisp","PctPolicAsian","PctPolicMinor","OfficAssgnDrugUnits","NumKindsDrugSeiz","PolicAveOTWorked","LandArea","PopDens","PctUsePubTrans","PolicCars","PolicOperBudg","LemasPctPolicOnPatr","LemasGangUnitDeploy","LemasPctOfficDrugUn","PolicBudgPerPop","ViolentCrimesPerPop"))
communities <- transform(communities,
                         State == as.integer(State),
                         County == as.integer(County),
                         Community == as.integer(Community))
communities$Region <- as.factor(ifelse(communities$State%in%c(8,22,24,32,34,35,42,44,51), "Northeast", ifelse(communities$State%in%c(16,17,18,19,25,26,28,30,37,39,46,55), "Midwest", ifelse(communities$State%in%c(1,5,9,10,11,12,20,21,23,27,36,40,45,47,48,52,54), "South", ifelse(communities$State%in%c(2,4,6,7,14,15,29,31,33,41,50,53,56), "West", "Territory")))))
communities <- relocate(communities, Region, .before="Fold")
communities <- transform(communities, Region = as.factor(Region))
head(communities)
view(communities)
```

Things to investigate:
--* RacePerCap & VCPP--
* Incomes/Poverty & VCPP
* Education & VCPP
* Employment & VCPP
* Rent & VCPP
* PolicPerPop & VCPP

Significant = 95% CIs have no overlap at VCPP = 0 and VCPP = 1
Not Sig = 95% CIs have some overlap at VCPP = 0 and VCPP = 1
PC = Per Capita

```{r racepercap data wrangling}
communities_racepercap <- communities |>
  pivot_longer(
    cols = c("WhitePerCap","BlackPerCap","IndianPerCap","AsianPerCap","OtherPerCap","HispPerCap"),
    names_to = "Race",
    values_to = "RPC_VCPP"
  )
ggplot(
  data = communities_racepercap,
  mapping = aes(x = Race, y = RPC_VCPP)) +
  geom_boxplot() +
  labs(
    title = "Race PC & Violent Crime",
    subtitle = "1990 & 1995 Data",
    x = "Race",
    y = "Violent Crime"
  )
```

```{r racepercap graph(s), echo = FALSE}
ggplot(data = communities,
       mapping = aes(
         x = WhitePerCap,
         y = ViolentCrimesPerPop
       )) +
  geom_boxplot()
```

```{r, incomes/poverty graph(s), echo = FALSE}
ggplot(data = communities,
       mapping = aes(
         x = MedIncome,
         y = ViolentCrimesPerPop
       )) +
  geom_point(alpha = 0.3, colour = "pink4") +
  geom_smooth(method = lm, fullrange = TRUE, colour = "pink1") +
  labs(title = "Med Income and Violent Crime",
       subtitle = "1990 & 1995 Data",
       x = "Med Income",
       y = "Violent Crime")
ggplot(data = communities,
       mapping = aes(
         x = MedFamInc,
         y = ViolentCrimesPerPop
       )) +
  geom_point(alpha = 0.3, colour = "pink4") +
  geom_smooth(method = lm, fullrange = TRUE, colour = "pink1") +
  labs(title = "Med Family Income and Violent Crime",
       subtitle = "1990 & 1995 Data",
       x = "Med Family Income",
       y = "Violent Crime")
ggplot(data = communities,
       mapping = aes(
         x = PerCapInc,
         y = ViolentCrimesPerPop
       )) +
  geom_point(alpha = 0.3, colour = "pink4") +
  geom_smooth(method = lm, fullrange = TRUE, colour = "pink1") +
  labs(title = "PC Income and Violent Crime",
       subtitle = "1990 & 1995 Data",
       x = "Per Capita Income",
       y = "Violent Crime")
ggplot(data = communities,
       mapping = aes(
         x = PctPopUnderPov,
         y = ViolentCrimesPerPop
       )) +
  geom_point(alpha = 0.3, colour = "pink4") +
  geom_smooth(method = lm, fullrange = TRUE, colour = "pink1") +
  labs(title = "Poverty and Violent Crime",
       subtitle = "1990 & 1995 Data",
       x = "% People in Poverty",
       y = "Violent Crime")
```
MedIncome - Significant
MedFamInc - Significant
PerCapInc - Significant
PctPopUnderPov - Significant

```{r, education graphs, echo = FALSE}
ggplot(data = communities,
       mapping = aes(
         x = PctLess9thGrade,
         y = ViolentCrimesPerPop
       )) +
  geom_point(alpha = 0.3, colour = "pink4") +
  geom_smooth(method = lm, fullrange = TRUE, colour = "pink1") +
  labs(title = "No HS Education and Violent Crime",
       subtitle = "1990 & 1995 Data",
       x = "% No HS Education",
       y = "Violent Crime")
ggplot(data = communities,
       mapping = aes(
         x = PctNotHSGrad,
         y = ViolentCrimesPerPop
       )) +
  geom_point(alpha = 0.3, colour = "pink4") +
  geom_smooth(method = lm, fullrange = TRUE, colour = "pink1") +
  labs(title = "No Diploma and Violent Crime",
       subtitle = "1990 & 1995 Data",
       x = "% No HS Diploma",
       y = "Violent Crime")
ggplot(data = communities,
       mapping = aes(
         x = PctBSorMore,
         y = ViolentCrimesPerPop
       )) +
  geom_point(alpha = 0.3, colour = "pink4") +
  geom_smooth(method = lm, fullrange = TRUE, colour = "pink1") +
  labs(title = "College Graduates and Violent Crime",
       subtitle = "1990 & 1995 Data",
       x = "% College Graduates",
       y = "Violent Crime")
```
PctLess9thGrade - Significant
PctNotHSGrad - Significant
PctBSorMore - Significant

```{r, employment graphs, echo = FALSE}
ggplot(data = communities,
       mapping = aes(
         x = PctEmploy,
         y = ViolentCrimesPerPop
       )) +
  geom_point(alpha = 0.3, colour = "pink4") +
  geom_smooth(method = lm, fullrange = TRUE, colour = "pink1") +
  labs(title = "% Employment and Violent Crime",
       subtitle = "1990 & 1995 Data",
       x = "% Employment",
       y = "Violent Crime")
ggplot(data = communities,
       mapping = aes(
         x = PctEmplManu,
         y = ViolentCrimesPerPop
       )) +
  geom_point(alpha = 0.3, colour = "pink4") +
  geom_smooth(method = lm, fullrange = TRUE, colour = "pink1") +
  labs(title = "% Manual Workers and Violent Crime",
       subtitle = "1990 & 1995 Data",
       x = "% Manual Workers",
       y = "Violent Crime")
ggplot(data = communities,
       mapping = aes(
         x = PctEmplProfServ,
         y = ViolentCrimesPerPop
       )) +
  geom_point(alpha = 0.3, colour = "pink4") +
  geom_smooth(method = lm, fullrange = TRUE, colour = "pink1") +
  labs(title = "% Professional Servicers and Violent Crime",
       subtitle = "1990 & 1995 Data",
       x = "% Professional Servicers",
       y = "Violent Crime")
ggplot(data = communities,
       mapping = aes(
         x = PctOccupMgmtProf,
         y = ViolentCrimesPerPop
       )) +
  geom_point(alpha = 0.3, colour = "pink4") +
  geom_smooth(method = lm, fullrange = TRUE, colour = "pink1") +
  labs(title = "% Management and Professional Workers and Violent Crime",
       subtitle = "1990 & 1995 Data",
       x = "% Management and Professional Workers",
       y = "Violent Crime")
```
PctEmploy - Significant
PctEmplManu - NOT Sig
PctEmplProfServ - Significant
PctOccupMgmtProf - Significant

```{r, housing graphs, echo = FALSE}
ggplot(data = communities,
       mapping = aes(
         x = MedRentPctHousInc,
         y = ViolentCrimesPerPop
       )) +
  geom_point(alpha = 0.3, colour = "pink4") +
  geom_smooth(method = lm, fullrange = TRUE, colour = "pink1") +
  labs(title = "Rent, Income, and Violent Crime",
       subtitle = "1990 & 1995 Data",
       x = "Med Rent as % of House Income",
       y = "Violent Crime")
ggplot(data = communities,
       mapping = aes(
         x = MedOwnCostPctInc,
         y = ViolentCrimesPerPop
       )) +
  geom_point(alpha = 0.3, colour = "pink4") +
  geom_smooth(method = lm, fullrange = TRUE, colour = "pink1") +
  labs(title = "Mortages, Income, and Violent Crime",
       subtitle = "1990 & 1995 Data",
       x = "Med Owners Cost as % of Income",
       y = "Violent Crime")
ggplot(data = communities,
       mapping = aes(
         x = MedOwnCostPctIncNoMtg,
         y = ViolentCrimesPerPop
       )) +
  geom_point(alpha = 0.1, colour = "pink4") +
  geom_smooth(method = lm, fullrange = TRUE, colour = "pink1") +
  labs(title = "Owners Costs, Income, and Violent Crime",
       subtitle = "1990 & 1995 Data",
       x = "Med Non-Mortgage Owners Cost as % of Income",
       y = "Violent Crime")
```
MedRentPctHousInc - Significant
MedOwnCostPctInc - Significant
MedOwnCostPctIncNoMtg - Significant
# i wish i had the variable of like, (numinshelters+numinstreet)/(population) but because it's all normalized htat's not possible so, no homeless graphs

```{r, police graphs, echo = FALSE}
ggplot(data = communities,
       mapping = aes(
         x = LemasSwFTPerPop,
         y = ViolentCrimesPerPop
       )) +
  geom_point(alpha = 0.3, colour = "pink4") +
  geom_smooth(method = lm, fullrange = TRUE, colour = "pink1") +
  labs(title = "Police and Violent Crime",
       subtitle = "1990 & 1995 Data",
       x = "Full-Time Officers Per Pop",
       y = "Violent Crime")
ggplot(data = communities,
       mapping = aes(
         x = LemasTotReqPerPop,
         y = ViolentCrimesPerPop
       )) +
  geom_point(alpha = 0.3, colour = "pink4") +
  geom_smooth(method = lm, fullrange = TRUE, colour = "pink1") +
  labs(title = "Police Requests and Violent Crime",
       subtitle = "1990 & 1995 Data",
       x = "Police Requests Per Pop",
       y = "Violent Crime")
ggplot(data = communities,
       mapping = aes(
         x = RacialMatchCommPol,
         y = ViolentCrimesPerPop
       )) +
  geom_point(alpha = 0.3, colour = "pink4") +
  geom_smooth(method = lm, fullrange = TRUE, colour = "pink1") +
  labs(title = "Police Demographics and Violent Crime",
       subtitle = "1990 & 1995 Data",
       x = "Community/Police Racial Match",
       y = "Violent Crime")
ggplot(data = communities,
       mapping = aes(
         x = PolicBudgPerPop,
         y = ViolentCrimesPerPop
       )) +
  geom_point(alpha = 0.3, colour = "pink4") +
  geom_smooth(method = lm, fullrange = TRUE, colour = "pink1") +
  labs(title = "Police Budget and Violent Crime",
       subtitle = "1990 & 1995 Data",
       x = "Police Budget Per Pop",
       y = "Violent Crime")
```
LemasSwFTPerPop - Significant
LemasTotReqPerPop - Significant
RacialMatchCommPol - Significant
PolicBudgPerPop - NOT Sig

```{r, regions & frequency graphs, echo = FALSE}
ggplot(data = communities,
       mapping = aes(
         x = ViolentCrimesPerPop,
         y = forcats::fct_reorder(Region, ViolentCrimesPerPop, .fun = median)
       )) +
  geom_boxplot() +
  labs(
    x = "Violent Crime",
    y = "US Region",
    title = "Violent Crime in the USA",
    subtitle = "1990"
  )
ggplot(data = communities,
       mapping = aes(x = ViolentCrimesPerPop)) +
  geom_density(fill = "pink4") +
  labs(
    x = "Violent Crime",
    y = "Frequency",
    title = "Violent Crime in the USA",
    subtitle = "1990 & 1995 Data"
  ) +
  theme_bw()
```

```{r additional/inspired by others}
communities |>
  filter(racePctWhite>=0.7) |> 
  ggplot(aes(x=PctLess9thGrade, y=PctSpeakEnglOnly)) +
  geom_point(alpha=0.3) +
  geom_smooth(method = lm)

# % NA values per column
communities %>% summarize_all(funs(sum(is.na(.)) / length(.)))
```

```{r unused/scrap code}
communities$Region <- communities |> mutate(State = case_when(
  communities$State%in%c(8,22,24,32,34,35,42,44,51) ~ "Northeast",
  communities$State%in%c(16,17,18,19,25,26,28,30,37,39,46,55) ~ "Midwest",
  communities$State%in%c(1,5,9,10,11,12,20,21,23,27,36,40,45,47,48,52,54) ~ "South",
  communities$State%in%c(2,4,6,7,14,15,29,31,33,41,50,53,56) ~ "West",
  TRUE ~ "Territory"
))
us.region <- function(State) {
  if (State %in% c(8,22,24,32,34,35,42,44,51)) {
    return("Northeast")
  } else if (State %in% c(16,17,18,19,25,26,28,30,37,39,46,55)) {
    return("Midwest")
  } else if (State %in% c(1,5,9,10,11,12,20,21,23,27,36,40,45,47,48,52,54)) {
    return("South")
  } else if (State %in% c(2,4,6,7,14,15,29,31,33,41,50,53,56)) {
    return("West")
  } else {
    return("Territory")
  }
}
communities <- communities |>
  mutate(USRegion = sapply(State, us.region))
ggplot(data = communities,
       mapping = aes(x = ViolentCrimesPerPop)) +
  geom_histogram(binwidth = 0.05) +
  labs(
    x = "Violent Crimes Per Population",
    y = "Frequency",
    title = "Violent Crime in the USA",
    subtitle = "1990 & 1995 Data"
  )
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
