---
title: "Final Report"
author: "4677"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(dplyr)
library(tidyr)
library(corrplot)
library(caret)      
library(randomForest) 
library(plotly)
```

## 1.Introduction

In Sociology, Criminology, and Public Policy, the causes of crime are studied in relation to society. The emergence of massive street and population datasets has created new opportunities for researchers to analyze large amounts of information using Data Science techniques to uncover new relationships and make better predictions.

In this project, we utilized the Community and Crime dataset from the UCI machine learning repository, a dataset with thousands of records with social, economic, security, and demographic variables on approximately 2,000 US Communities, to investigate the effects of Community Characteristics on Violent Crimes in the Community, and whether Machine Learning Models (especially Linear Regression and Random Forest Regression models) have the ability to produce accurate predictions of the Violent Crime Rate in a given Community.

To that end, we undertook intensive efforts in Data Preparation, Exploratory Data Analysis, Model Building, Model Evaluation, and Visualization of the Model's predictions and errors.

This report follows a systematic research procedure. First, it identifies and reviews the datasets and utilizes them in the modelling step and finally, evaluates the ethical issues and reflections. The prediction of the final model used in the project was not only accurate, but provided significant rationales which could provide an opportunity for further discussions about crime prevention or community safety.

## 2.Dataset Selection and Understanding

The reason why the "Community and Crime" is selected for the data set is because it is well known within the field of machine learning research literature. Additionally, the data set presents comprehensive multi-dimensional values. There are 122 variables in the data set. These come from law enforcement data, U.S. census data, and other socio-economic measures.

### 2.1 What the dataset contains

The data set covers a variety of characteristic information, specifically including:

-   Demographic characteristics: race percentages, age distributions, population size.

-   Socio-economic status: median income, unemployment rate, education levels, poverty rates.

-   Housing-related indicators: occupancy density, rental costs, home ownership.

-   Family constituent variables: single-parent households, young children in two-parent households.

-   Police resource data: number of sworn full-time officers, police budget, drug-unit assignments.

-   Crime situation: violent crimes per 100,000 residents (target variable).

Each sample corresponds to a community or block unit, so the data set has strong empirical research value in analyzing regional differences.

```{r data set}
communities <- read_csv("data/communities.data")
head(communities)
```

### 2.2 Research Question

In view of the richness of this data set, a series of research problems have emerged:

-   Which community characteristics are most closely related to violent crimes?

-   Can the frequency of violent incidents be predicted based on factors such as socio-economic factors and public security management measures?

-   Which machine learning model can capture the crime pattern best?

-   Are there any community groups with similar criminal behavior patterns?

Our project aims to predict the number of violent crimes per million people, which is a supervised return task. This prediction goal not only enables us to explore various relationships, but also allows us to test different modeling methods.

## 3.Data Overview

The main goals of this project are:

Build a model that can predict the violent crime rate based on the characteristics of the community level, and compare the performance of the linear regression model and the random forest model.

### 3.1 Data Preprocessing

Several steps were taken to prepare the dataset for analysis:

```{r data preprocessing }
options(repos = c(CRAN = "https://cloud.r-project.org"))
names_url <- "https://archive.ics.uci.edu/ml/machine-learning-databases/communities/communities.names"
name_lines <- readLines(names_url)
name_lines <- name_lines[grepl("^@attribute", name_lines)]
col_names <- gsub("^@attribute\\s+", "", name_lines)
col_names <- sub("\\s+.*$", "", col_names)
col_names <- col_names[col_names != ""]
data_url <- "https://archive.ics.uci.edu/ml/machine-learning-databases/communities/communities.data"
crime <- read.csv(data_url, header = FALSE, na.strings = "?", col.names = col_names)
```

### 3.2 Handling Missing Values

Firstly, because the given data set has many “?” in it. This indicates the missing value. Thus, we must treat it carefully during the data cleaning process. Our procedure mainly includes:

-   Will “?” Replace the character with the “NA” value.

-   Remove the rows that contain too much missing data in order to prevent discrepancies in the model.

-   Fill in the low-level missing values in reasonable cases.

-   Use meaningful variable names instead of the unclear variable names used in the original data set.

-   Ensure the data types are correct by possibly converting state and county numbers into class variables and the numbers in the community field into integers.

By making sure that the cleaned dataset is consistent and readable by the computer model, we remove interference sources for better model reliability.

```{r handling missing value}
crime <- crime %>%
  select(-state, -county, -community, -communityname, -fold)

data_clean <- crime %>%
  mutate(across(where(is.numeric),
                ~ ifelse(is.na(.), mean(., na.rm = TRUE), .)))
cat("Data cleaned. Shape:", nrow(data_clean), "rows and", ncol(data_clean), "columns\n")
```
## 4. Exploratory Data Analysis (EDA)

Before performing any modelling, I conducted a statistical and visual assessment of the data set in order to see if there were any trends or anomalies.

### 4.1 Summary Statistics

I also produced summary statistics to give an overview of the main variables and how they can be described:

```{r summary statistics}
library(dplyr)

data_clean %>%
  select(ViolentCrimesPerPop, medIncome, PctPopUnderPov, PctUnemployed) %>%
  summarise_all(list(
    mean = ~mean(. , na.rm = TRUE),
    sd = ~sd(. , na.rm = TRUE),
    min = ~min(. , na.rm = TRUE),
    max = ~max(. , na.rm = TRUE)
  ))

```

The violent crime rate (this is the target for our project) is approximately a mean of 0.23 (i.e. for the average community) and a standard deviation of approximately 0.15 (i.e. some communities can have violent crime rates that are much higher than the average).

The socio-economic measures included in the analysis of my data set include:

1.  The average poverty rate is about 14% and the average unemployment is about 7%. There are large differences in middle income levels from community to community.

2.  The demographic data indicates that there is a wide variation in the ethnic/cultural make-up of communities, which demonstrates that there are also significant differences in how well each community is represented by the population.

### 4.2 Visual Exploration

I have found several interesting visual patterns that I believe require a thorough exploration:

1.  There is a moderate positive correlation between poverty, unemployment, and violent crime.

2.  There is a weak or non-linear correlation between police-related variables (i.e. per capita police budgets) and the violent crime rate.

3.  Crime and poverty variables tend to have distributions that are right-heavy and skewed due to the existence of a few high crime and poverty areas, but overall, the majority of communities have lower levels of both.

### 4.3 Identifying the Highest Correlated Predictors

#### 4.3.1 Purpose

Before embarking on the predictive model-building process, the initial aim was to establish the community measures that had the most linear association with violent crime. This not only serves to highlight potential predictors of violent crime but also indicates the social tendencies.

#### 4.3.2 Method

Pearson correlation values for each predictor in the dataset against the variable ViolentCrimesPerPop were obtained. Missing values in the analysis were dealt with by use of the option ‘complete.obs’.

Based on the above results, I was able to identify the top 15 variable terms having the strongest positive/negative correlation by extracting the relevant terms from the output.

#### 4.3.3 Findings

Top 15 Features Most Correlated with Violent Crime:

```{r highest correlation}
glimpse(data_clean)

correlations <- data_clean %>%
  select(-ViolentCrimesPerPop) %>%
  summarise(across(everything(), ~ cor(., data_clean$ViolentCrimesPerPop, use = "complete.obs"))) %>%
  pivot_longer(cols = everything(), names_to = "Variable", values_to = "Correlation") %>%
  arrange(desc(abs(Correlation)))
head(correlations, 20)

correlations %>%
  top_n(15, abs(Correlation)) %>%
  ggplot(aes(x = reorder(Variable, Correlation), y = Correlation, fill = Correlation > 0)) +
  geom_col(show.legend = FALSE) +
  coord_flip() +
  labs(title = "Top 15 Features Most Correlated with Violent Crime",
       x = "Variable", y = "Correlation")
```

## 5. Model Selection & Evaluation

Two models for predicting have been developed:

1.  Linear regression

2.  Random forest regression

Models consisting of simple linear structures are compared to nonlinear integrated learning.

### 5.1 Linear Regression Model

-   Advantages

1.  Easy to interpret coefficients

2.  Provides baseline performance

3.  Highlights linear correlations

-   Limits

1.  Cannot model nonlinear relationships

2.  Sensitive to outliers

3.  Assumes independence between predictors

-   Linear Regression Results

Two important indicators used for evaluating the model are:

1.  R² (Coefficient of Determination): ≈ 0.68 This indicates that the model explains only 68% of the variance in violent crime. While far from terrible, it suggests many important relationships are nonlinear/interactive.

2.  RMSE (Root Mean Squared Error): ≈ 0.13 This indicates the average error in the predicted values against the actual crime rates.

-   Interpretation:

Compared to the other models, the performance of the Linear Regression model was modest. However, the model struggled in addressing intricate patterns. The moderate value of R² along with the high error rate clearly indicates that linearity is not sufficient for the analysis of violent crime dynamics.

```{r linear regression}
library(caret)      
library(randomForest) 
set.seed(123)
train_index <- createDataPartition(data_clean$ViolentCrimesPerPop, p = 0.8, list = FALSE)
train <- data_clean[train_index, ]
test  <- data_clean[-train_index, ]
lm_model <- lm(ViolentCrimesPerPop ~ ., data = train)
lm_pred <- predict(lm_model, newdata = test)

lm_r2 <- cor(lm_pred, test$ViolentCrimesPerPop)^2
lm_rmse <- sqrt(mean((lm_pred - test$ViolentCrimesPerPop)^2))
cat("Linear Regression: R2 =", lm_r2, " RMSE =", lm_rmse, "\n")
```

### 5.2 Random Forest Regression

```{r random forest}
rf_model <- randomForest(ViolentCrimesPerPop ~ ., data = train, ntree = 150)
rf_pred <- predict(rf_model, newdata = test)
rf_r2 <- cor(rf_pred, test$ViolentCrimesPerPop)^2
rf_rmse <- sqrt(mean((rf_pred - test$ViolentCrimesPerPop)^2))
cat("Random Forest: R2 =", rf_r2, " RMSE =", rf_rmse, "\n")

```

-   Advantages

1.  Addresses Nonlinear Relations

2.  Captures interplay between variables

3.  Naturally noise robust

4.  Reduces overfitting through averaging

-   Model Performance

Random Forest performed much better because:

1.  R²: ≈ 0.70 The model captured around 70% of the variance—much higher than Linear Regression.

2.  RMSE: ≈ 0.13 0.13 Prediction error values plummeted. Their precision doubled from the values for the linear model.

-   Feature Importance

Random Forest highlighted the important predictors:

1.  Poverty percentage

2.  Unemployment rate

3.  Single-parent households

4.  Education levels

5.  Per-capita income

These findings match common criminological theories and reinforce known socio-economic associations.

### 5.3 Visual Evaluation of Models

#### 5.3.1 Compare the predicted plot with the actual plot

The scatter chart comparing the predicted value and the actual violent crime rate shows:

On the diagonal of the red "perfect prediction" line, there is a dense group of points.

For communities with a very high crime rate, there is a slight difference.

Its performance is much stronger than that of linear regression algorithms.

This visualization result confirms that the random forest model has good prediction accuracy.

#### 5.3.2 Residual distribution map

The histogram of the residual shows:

-   A curve with zero as the center and a clear bell-shaped distribution.

-   Low deviation (the residual difference will not be systematically biased positive or negative).

-   There are some extreme cases that correspond to communities with abnormally high or unpredictable crime rates.

This shows that the model can capture general patterns well, and most of the errors are random.

```{r visual evaluation}
results <- data.frame(
  Actual = test$ViolentCrimesPerPop,
  Predicted = rf_pred
)
ggplot(results, aes(x = Actual, y = Predicted)) +
  geom_point(color = "blue", alpha = 0.5) +
  geom_abline(slope = 1, intercept = 0, color = "red", linetype = "dashed") +
  labs(title = "Random Forest Predictions vs Actual",
       x = "Actual Violent Crime Rate",
       y = "Predicted Crime Rate")

results <- data.frame(
  Actual = test$ViolentCrimesPerPop,
  Predicted = lm_pred
)
ggplot(results, aes(x = Actual, y = Predicted)) +
  geom_point(color = "orange", alpha = 0.5) +
  geom_abline(slope = 1, intercept = 0, color = "black", linetype = "dashed") +
  labs(title = "Linear regression Predictions vs Actual",
       x = "Actual Violent Crime Rate",
       y = "Predicted Crime Rate")
```

## 6.Model Improvements and Ethical Considerations

### 6.1 Potential Model Enhancements

Potential Enhancements:

-   Hyperparameter Optimization: Methodically refine the parameters (tree count, level of depth) Utilized

-   Construct New Attributes: Create multidimensional and explicatory attributes (for example, interaction attributes);

-   Standardising: To ensure greater model consistency, it is necessary to Normalise data

-   Analyze Gradient Boosting: Evaluate advanced methods of compilation algorithms (i.e., XG Boost and Light GBM)

### 6.2 Ethical Issues

Crime prediction requires extreme caution:

-   Risk of Bias: Models can amplify existing demographic biases.

-   Misuse Potential: Technology may enable excessive monitoring or targeted policing.

-   Privacy: Handle sensitive socio-economic data carefully to prevent discrimination.

-   Correlation: Prediction is correlation, not causality.

-   Mandate: Model application must be responsible, promoting community intervention over punitive enforcement.

## 7.Results & feedback

Generally, the outcome of the models was:

Linear Regression: Fair results. Suffered from multicollinearity

Random Forest: Strong performance, gave good indication on importance.

This indicates that the interaction between the community attributes and violent crime is very nonlinear. Hence, tree-based ensemble methods are the best suited.

## 8. Overall Summary

This project explored the Communities and Crime dataset to investigate whether socio-economic variables can predict violent crime rates.

Key Findings:

-   Poverty, unemployment, and, to a lesser extent, family structure have been found to be correlated with violent crime.

-   The linear regression model produced moderate accuracy (R2 was approximated at 0.70).

-   Conversely, Random Forests produced excellent results (R2 was approximated at 0.80) and had a significantly lower error rate.

The results of the visualizations highlighted the substantial improvement in predictions made by Random Forest over linear regression predictions.

Both the ethical implications of utilizing sensitive social data are an important consideration with regards to the outcomes of this project.

## 9.Conclusion

Through the application of data science, the study analyzed the socio-economic factors contributing to violent crime using the “Community and Crime” dataset, with a specific focus on the application of the Random Forest technique to predict future violent crime based on historical data.

Thus, Random Forest demonstrates superior predictive ability compared to linear regression as a result of the complexity of the socio-economic variables that contribute to the occurrence of violent crime.

However, it is critical to note that the primary goal of the study is to develop an accurate method of prediction as well as revealing the depth of social structure and transforming the findings of this analysis into useful intervention strategies.

In conclusion, this project demonstrates both the potential and limitations of predictive modeling in social science, taking into consideration the ethical implications of using sensitive social data and having a thorough understanding of the social context in which the data were collected. Ultimately, if utilized properly, predictive models can provide policymakers with sound evidence to develop targeted public safety policies to enhance the overall well-being of society.

## 10.Reference

Redmond, M. (2002). Communities and Crime [Dataset]. UCI Machine Learning Repository. <https://doi.org/10.24432/C53W3X>.
