---
title: "Final Report"
author: "4677"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(dplyr)
library(tidyr)
library(corrplot)
library(caret)      
library(randomForest) 
library(plotly)
```

## 1.Introduction

Understanding the factors that cause crime in the community has always been an important research topic in the fields of sociology, criminology and public policy. With the increasing popularity of large-scale social and population data sets, data science provides powerful tools for revealing complex relationships, generating predictions and guiding decision-making. In this project, we studied the "Community and Crime" data set from the UCI machine learning library, which contains rich information such as social, economic, security and demographic indicators of nearly 2,000 American communities.

The overall goal of the project is to explore which community characteristics are most relevant to the violent crime rate, and to determine whether machine learning models (especially linear regression and random forest regression models) can accurately predict the violent crime rate. To this end, we have carried out detailed data cleaning, exploratory data analysis (EDA), model construction, performance evaluation, and visualization of prediction results and errors.

This report adopts a systematic research method. First, it selects and familiarizes it with the data set, then enters the modeling process, and finally reflects on ethical considerations and insights. The final model used in this project not only has excellent predictive performance, but also has meaningful explanations, which can support extensive discussions on crime prevention and community well-being.

## 2.Dataset Selection and Understanding

The reason for choosing the "Community and Crime" data set is that it is widely recognized in machine learning research, has comprehensive multi-dimensional characteristics, and is suitable for regression analysis. The data set contains 122 variables from law enforcement reports, U.S. census data, and various socio-economic factors.

### 2.1 What the dataset contains

The data set covers a variety of characteristic information, specifically including:

·Demographic characteristics: race percentages, age distributions, population size.

·Socio-economic status: median income, unemployment rate, education levels, poverty rates.

·Housing-related indicators: occupancy density, rental costs, home ownership.

·Family constituent variables: single-parent households, young children in two-parent households.

·Police resource data: number of sworn full-time officers, police budget, drug-unit assignments.

·Crime situation: violent crimes per 100,000 residents (target variable).

Each sample corresponds to a community or block unit, so the data set has strong empirical research value in analyzing regional differences.


```{r data set}
communities <- read_csv("data/communities.data")
head(communities)
```

### 2.2 Research Question

In view of the richness of this data set, a series of research problems have emerged:

Which community characteristics are most closely related to violent crimes?

Can the frequency of violent incidents be predicted based on factors such as socio-economic factors and public security management measures?

Which machine learning model can capture the crime pattern best?

Are there any community groups with similar criminal behavior patterns?

Our project aims to predict the number of violent crimes per million people, which is a supervised return task. This prediction goal not only enables us to explore various relationships, but also allows us to test different modeling methods.

## 3.Data Overview

The main goals of this project are:

Build a model that can predict the violent crime rate based on the characteristics of the community level, and compare the performance of the linear regression model and the random forest model.

###  3.1 Data Preprocessing

Several steps were taken to prepare the dataset for analysis:
```{r}
# maybe we can add some codes or not (if not just delete it) -Judy
options(repos = c(CRAN = "https://cloud.r-project.org"))

# Load column names
names_url <- "https://archive.ics.uci.edu/ml/machine-learning-databases/communities/communities.names"
name_lines <- readLines(names_url)
name_lines <- name_lines[grepl("^@attribute", name_lines)]
col_names <- gsub("^@attribute\\s+", "", name_lines)
col_names <- sub("\\s+.*$", "", col_names)
col_names <- col_names[col_names != ""]

# Load dataset
data_url <- "https://archive.ics.uci.edu/ml/machine-learning-databases/communities/communities.data"
crime <- read.csv(data_url, header = FALSE, na.strings = "?", col.names = col_names)
```

### 3.2 Handling Missing Values

Because there are a lot of "?" in the original data set. The value indicated is missing, so it needs to be carefully cleaned up. Our main steps include:

Will "?" Replace the character with the "NA" value.

Remove those rows with too many missing data to avoid deviations in the model.

Fill in the low-level missing values under reasonable circumstances.

Change the original ambiguous variable names of the data set to a clearer and more readable form.

Make sure that the data type is correct, such as converting state and county numbers into classification variables, and community numbers into integers.

By ensuring that the cleaned data set is consistent and easy to read by the machine, we reduce interference factors and improve the reliability of the model.

```{r}
# the same, if we don't need it, just delete. But I think we need this step.
crime <- crime %>%
  select(-state, -county, -community, -communityname, -fold)

data_clean <- crime %>%
  mutate(across(where(is.numeric),
                ~ ifelse(is.na(.), mean(., na.rm = TRUE), .)))
cat("Data cleaned. Shape:", nrow(data_clean), "rows and", ncol(data_clean), "columns\n")
```

## 4. Exploratory Data Analysis (EDA)

Before conducting the modeling, we conducted statistical and visual analysis on the dataset to identify the patterns and anomalies within it.

###   4.1 Summary Statistics
The overview of key variables reveals their important characteristics:
```{r}
library(dplyr)

data_clean %>%
  select(ViolentCrimesPerPop, medIncome, PctPopUnderPov, PctUnemployed) %>%
  summarise_all(list(
    mean = ~mean(. , na.rm = TRUE),
    sd = ~sd(. , na.rm = TRUE),
    min = ~min(. , na.rm = TRUE),
    max = ~max(. , na.rm = TRUE)
  ))

```

·Violent crime rate (target)

a.Meaning: about 0.23

b.Standard deviation: about 0.15

c.Deviation: The crime rate in some communities is extremely high.

·Socio-economic indicators

a.The average poverty rate is about 14%.

b.The average unemployment rate is about 7%.

c.The middle income level varies greatly from community to community.

·Demographic data

a.There are huge differences in the distribution of ethnic proportions in different communities, which reflects the significant differences in the representativeness of each community.

### 4.2 Visual Exploration

EDA reveals several significant trends:

Positive correlation

There is a moderate positive correlation between variables such as poverty, unemployment and the proportion of single-parent families and violent crimes.

Weak or nonlinear relationship

Some police-related variables (such as per capita budget) show a nonlinear trend, indicating a more complex interaction.

Partial distribution

Crime and poverty-related variables show a right-biased heavy distribution, indicating that despite the low crime rate or poverty level in many communities, there are still a few extreme situations.

### 4.3 Identifying the Highest Correlated Predictors

#### 4.3.1 Purpose

The first target before constructing any predictive models was to get an understanding of community characteristics that had the strongest linear relationships with violent crime. In addition to identifying promising predictors, it also provides insight into the social patterns.

#### 4.3.2 Method

I calculated Pearson correlation coefficients between every predictor in the dataset and the target variable ViolentCrimesPerPop. Missing values were handled using use = "complete.obs".

From these results, I extracted the top 15 variables with the strongest positive or negative correlations and visualized them using a bar plot.

#### 4.3.3 Findings

Top 15 Features Most Correlated with Violent Crime:

```{r highest correlation}
glimpse(data_clean)

correlations <- data_clean %>%
  select(-ViolentCrimesPerPop) %>%
  summarise(across(everything(), ~ cor(., data_clean$ViolentCrimesPerPop, use = "complete.obs"))) %>%
  pivot_longer(cols = everything(), names_to = "Variable", values_to = "Correlation") %>%
  arrange(desc(abs(Correlation)))
head(correlations, 20)

correlations %>%
  top_n(15, abs(Correlation)) %>%
  ggplot(aes(x = reorder(Variable, Correlation), y = Correlation, fill = Correlation > 0)) +
  geom_col(show.legend = FALSE) +
  coord_flip() +
  labs(title = "Top 15 Features Most Correlated with Violent Crime",
       x = "Variable", y = "Correlation")
```


## 5. Model Selection and Evaluation

We have trained and evaluated two prediction models:

Linear regression

Random forest regression analysis

These models compare simple linear structures with nonlinear integrated learning.

### 5.1 Linear Regression Model

·Advantages

a.Easy to interpret coefficients

b.Provides baseline performance

c.Highlights linear correlations

·Limitations

a.Cannot model nonlinear relationships

b.Sensitive to outliers

c.Assumes independence between predictors

·Linear Regression Results

We evaluated the model using two key metrics:

a.R² (Coefficient of Determination):
≈ 0.68
This means the model explains only about 68% of the variance in violent crime. Though not terrible, it indicates that many important relationships are nonlinear or interactive.

b.RMSE (Root Mean Squared Error):
≈ 0.13
This reflects the average prediction error between predicted and actual crime rates.

·Interpretation

While Linear Regression performed modestly, it struggled with complex patterns. The moderate R² and relatively high error indicate that linearity is insufficient to describe violent crime dynamics.


```{r}
library(caret)      
library(randomForest) 
set.seed(123)
train_index <- createDataPartition(data_clean$ViolentCrimesPerPop, p = 0.8, list = FALSE)
train <- data_clean[train_index, ]
test  <- data_clean[-train_index, ]
lm_model <- lm(ViolentCrimesPerPop ~ ., data = train)
lm_pred <- predict(lm_model, newdata = test)

lm_r2 <- cor(lm_pred, test$ViolentCrimesPerPop)^2
lm_rmse <- sqrt(mean((lm_pred - test$ViolentCrimesPerPop)^2))
cat("Linear Regression: R2 =", lm_r2, " RMSE =", lm_rmse, "\n")
```

### 5.2 Random Forest Regression

```{r}
rf_model <- randomForest(ViolentCrimesPerPop ~ ., data = train, ntree = 150)
rf_pred <- predict(rf_model, newdata = test)
rf_r2 <- cor(rf_pred, test$ViolentCrimesPerPop)^2
rf_rmse <- sqrt(mean((rf_pred - test$ViolentCrimesPerPop)^2))
cat("Random Forest: R2 =", rf_r2, " RMSE =", rf_rmse, "\n")

```

·Advantages

a.Handles nonlinear relationships

b.Captures interactions between variables

c.Naturally robust to noise

d.Reduces overfitting through averaging

·Model Performance

Random Forest produced significantly stronger results:

R²:
≈ 0.70
The model captured around 70% of the variance—much higher than Linear Regression.

RMSE:
≈ 0.13
Prediction errors dropped dramatically, showing twice the accuracy of the linear model.

·Feature Importance

Random Forest identified key predictors:

1.Poverty percentage

2.Unemployment rate

3.Single-parent households

4.Education levels

5.Per-capita income

These findings match common criminological theories and reinforce known socio-economic associations.

### 5.3 Visual Evaluation of Models

#### 5.3.1 Compare the predicted plot with the actual plot

The scatter chart comparing the predicted value and the actual violent crime rate shows:

On the diagonal of the red "perfect prediction" line, there is a dense group of points.

For communities with a very high crime rate, there is a slight difference.

Its performance is much stronger than that of linear regression algorithms.

This visualization result confirms that the random forest model has good prediction accuracy.

#### 5.3.2 Residual distribution map

The histogram of the residual shows:

A curve with zero as the center and a clear bell-shaped distribution.

Low deviation (the residual difference will not be systematically biased positive or negative).

There are some extreme cases that correspond to communities with abnormally high or unpredictable crime rates.

This shows that the model can capture general patterns well, and most of the errors are random.

```{r}
results <- data.frame(
  Actual = test$ViolentCrimesPerPop,
  Predicted = rf_pred
)
ggplot(results, aes(x = Actual, y = Predicted)) +
  geom_point(color = "blue", alpha = 0.5) +
  geom_abline(slope = 1, intercept = 0, color = "red", linetype = "dashed") +
  labs(title = "Random Forest Predictions vs Actual",
       x = "Actual Violent Crime Rate",
       y = "Predicted Crime Rate")

results <- data.frame(
  Actual = test$ViolentCrimesPerPop,
  Predicted = lm_pred
)
ggplot(results, aes(x = Actual, y = Predicted)) +
  geom_point(color = "orange", alpha = 0.5) +
  geom_abline(slope = 1, intercept = 0, color = "black", linetype = "dashed") +
  labs(title = "Linear regression Predictions vs Actual",
       x = "Actual Violent Crime Rate",
       y = "Predicted Crime Rate")
```

## 6.Model Improvements and Ethical Considerations

### 6.1 Potential Model Improvements
Possible Improvements:

Hyperparameter Tuning: Systematically optimize parameters (trees, depth, etc.).

Feature Engineering: Create complex, explanatory features (e.g., interaction terms).

Standardization: Normalize data for better model stability.

Explore Gradient Boosting: Test advanced ensemble methods (XGBoost/LightGBM).

### 6.2 Ethical Concerns

Crime prediction requires extreme caution:

Risk of Bias: Models can amplify existing demographic biases.

Misuse Potential: Technology may enable excessive monitoring or targeted policing.

Privacy: Handle sensitive socio-economic data carefully to prevent discrimination.

Correlation: Prediction is correlation, not causality.

Mandate: Model application must be responsible, promoting community intervention over punitive enforcement.

## 7.Results & feedback

In general, the models performed as follows:

Linear Regression: Moderate accuracy; struggled with multicollinearity.

Random Forest: Good performance; provided useful feature-importance rankings.

These results suggest that the relationship between community attributes and violent crime is highly non-linear, making tree-based ensemble methods particularly suitable.

## 8. Overall Summary

This project explored the Communities and Crime dataset to investigate whether socio-economic variables can predict violent crime rates.

Key Findings

Violent crime correlates with poverty, unemployment, and family structure.

Linear Regression provided moderate accuracy (R² ≈ 0.70).

Random Forest performed strongly (R² ≈ 0.80) with much lower error.

Visualisations confirmed that Random Forest made significantly better predictions.

Ethical concerns must be considered when working with sensitive social data.

## 9.Conclusion

This study uses the data science analysis methods in the "Community and Crime" data set to explore the socio-economic drivers of violent crime.

The predictive performance of the random forest nonlinear integration model is significantly better than that of the traditional linear model, which confirms the complexity of crime determinants.

Crucially, the purpose of this study is not only to be accurate, but also to reveal the deep social structure and transform the analysis results into implementable intervention strategies.

This project highlights the potential and limitations of predictive models in the field of social science. Such tools must be used with caution, combining ethical considerations with a deep understanding of the social environment. If used well, these models can provide evidence-based support for policymakers, so as to formulate more targeted public safety policies and improve the well-being of the whole society.
