---
title: "report"
author: "4677"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(tidymodels)
library(dplyr)
library(tidyr)
library(corrplot)
library(caret)      
library(randomForest) 
library(plotly)
```

## Data

This is the original data set.

The data set starts with variables on each community's state, county, and community ID and name. The State variable is listed as 1-56, with each U.S. community or territory in alphabetical order.
The data set then lists variables to do with populations, household size, race, ages, the size of urban populations, incomes, retirements, education, employment and field of employment, divorce, families, immigration, language, home size, home ownership, rent, homeless populations, and police statistics, ending with the variable measuring the amount of violent crime proportional to the population.

```{r data set}
communities <- read_csv("data/communities.data")
head(communities)
```

## Research Question

This is our research question.

This is why we chose this question.

## Data Setup

The values are normalized. This is what that means. These are some issues that arise from this.
One example of an issue that arises from the use of normalized variables comes from when we attempted to measure homeless populations. We wanted to combine the variables NumInShelters and NumStreet; however, since they were normalized, we could not combine them to compare the population to the ViolentCrimesPerPop variable. If the variables had not been normalized, this would have been possible.

There are variables that appear similar or the same but have different values. For example, BlackPerCap and RacePctBlack, have different values despite the fact that they are both normalized and measuring the same thing. In the first row, with community Lakewoodcity, RacePctBlack has a value of 0.02, while BlackPerCap has a value of 0.32. We have yet to construct a hypothesis as to why this might be true.

There are missing variables in the initial data set. This is because the initial data set was created using multiple sources, so not all of the sources had data for the complete set of communities. So, when placed together, many rows are missing observations.

This is how we cleaned the original data set for our use.
We cleaned the original data set by setting the names to make sure that they were standardized, and converting different columns into different types. State and County, for example, was converted into factors in order to allow communities to be grouped by location if necessary.

```{r data setup}
communities <- setNames(communities, c("State","County","Community","CommunityName","Fold","Population","HouseholdSize","RacePctBlack","RacePctWhite","RacePctAsian","RacePctHisp","AgePct12t21","AgePct12t29","AgePct16t24","AgePct65up","NumbUrban","PctUrban","MedIncome","PctWWage","PctWFarmSelf","PctWInvInc","PctWSocSec","PctWPubAsst","PctWRetire","MedFamInc","PerCapInc","WhitePerCap","BlackPerCap","IndianPerCap","AsianPerCap","OtherPerCap","HispPerCap","NumUnderPov","PctPopUnderPov","PctLess9thGrade","PctNotHSGrad","PctBSorMore","PctUnemployed","PctEmploy","PctEmplManu","PctEmplProfServ","PctOccupManu","PctOccupMgmtProf","MalePctDivorce","MalePctNevMarr","FemalePctDiv","TotalPctDiv","PersPerFam","PctFam2Par","PctKids2Par","PctYoungKids2Par","PctTeen2Par","PctWorkMomYoungKids","PctWorkMom","NumIlleg","PctIlleg","NumImmig","PctImmigRecent","PctImmigRec5","PctImmigRec8","PctImmigRec10","PctRecentImmig","PctRecImmig5","PctRecImmig8","PctRecImmig10","PctSpeakEnglOnly","PctNotSpeakEnglWell","PctLargHouseFam","PctLargHouseOccup","PersPerOccupHous","PersPerOwnOccHous","PersPerRentOccHous","PctPersOwnOccup","PctPersDenseHous","PctHousLess3BR","MedNumBR","HousVacant","PctHousOccup","PctHousOwnOcc","PctVacantBoarded","PctVacMore6Mos","MedYrHousBuilt","PctHousNoPhone","PctWOFullPlumb","OwnOccLowQuart","OwnOccMedVal","OwnOccHiQuart","RentLowQ","RentMedian","RentHighQ","MedRent","MedRentPctHousInc","MedOwnCostPctInc","MedOwnCostPctIncNoMtg","NumInShelters","NumStreet","PctForeignBorn","PctBornSameState","PctSameHouse85","PctSameCity85","PctSameState85","LemasSwornFT","LemasSwFTPerPop","LemasSwFTFieldOps","LemasSwFTFieldPerPop","LemasTotalReq","LemasTotReqPerPop","PolicRepPerOffic","PolicPerPop","RacialMatchCommPol","PctPolicWhite","PctPolicBlack","PctPolicHisp","PctPolicAsian","PctPolicMinor","OfficAssgnDrugUnits","NumKindsDrugSeiz","PolicAveOTWorked","LandArea","PopDens","PctUsePubTrans","PolicCars","PolicOperBudg","LemasPctPolicOnPatr","LemasGangUnitDeploy","LemasPctOfficDrugUn","PolicBudgPerPop","ViolentCrimesPerPop"))
communities <- transform(communities,
                         State == as.factor(State),
                         County == as.factor(County),
                         Community == as.integer(Community))
```

## randomForest

This is what randomForest does, with a brief explanation as to how, since it is not one of the packages we have discussed in class.

This is how we used randomForest in exploring our data.

```{r randomForest}
```

## Highest Correlation

These are the variables with the most significant correlation to the violent crime rate.

```{r highest correlation}
data <- communities %>%
  select(-c(State, County, Community, CommunityName, Fold))
colSums(is.na(data)) %>% sort(decreasing = TRUE)
data_clean <- data %>% select(where(~ mean(is.na(.)) < 0.3))
data_clean <- data_clean %>% mutate(across(everything(), ~ replace_na(., mean(., na.rm = TRUE))))
correlations <- data_clean %>%
  select(-ViolentCrimesPerPop) %>%
  summarise(across(everything(), ~ cor(., data_clean$ViolentCrimesPerPop, use = "complete.obs"))) %>%
  pivot_longer(cols = everything(), names_to = "Variable", values_to = "Correlation") %>%
  arrange(desc(abs(Correlation)))
correlations %>%
  top_n(15, abs(Correlation)) %>%
  ggplot(aes(x = reorder(Variable, Correlation), y = Correlation, fill = Correlation > 0)) +
  geom_col(show.legend = FALSE) +
  coord_flip() +
  labs(title = "Top 15 Features Most Correlated with Violent Crime",
       x = "Variable", y = "Correlation")
```

## Adjusted r-squared

These are the top 5 variables that produce the highest adjusted r-squared score.

```{r adjusted r-squared}
```

This is how we discovered the top 5 variables.

These are the scatterplot/linear model graphs of the top 5 variables.

```{r adjusted r-squared graphs, echo = FALSE}
```

## Possible explanations

There are many possible explanations for why certain variables are more or less correlated with the violent crime rate. 
PctIlleg likely has a high correlation due to the lack of legal work opportunities for people who are not in the United States on a Visa, Green Card, or who are citizens, due to many employeres requiring things like SSNs and proof of visas. As such, many illegal immigrants may be forced to turn to crime, creating a larger crime community, which can lead to an increased prevalence of violence due to increases in power.

Examples: Historical racism, racial discrepencies, slavery, and inequality, poverty and desperation, reasons for immigration, privilege

Note that the `echo = FALSE` parameter can be added to the code chunk to prevent printing of the R code that generated the plot.