---
title: "report"
author: "4677"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(dplyr)
library(tidyr)
library(corrplot)
library(caret)      
library(randomForest) 
library(plotly)
```

## Data

This is the original data set.

```{r data set}
communities <- read_csv("data/communities.data")
```

## Research Question

This is our research question.

This is why we chose this question.

## Data Setup

The values are normalized. This is what that means. These are some issues that arise from this.

There are varaibles that appear similar or the same but have different values. These are our hypotheses as to why this is true.

There are missing variables in the initial data set. This is why.

This is how we cleaned the original data set for our use.

## randomForest

This is what randomForest does, with a brief explanation as to how.

This is how we used randomForest in exploring our data.

## Highest Correlation

These are the variables with the most significant correlation to the violent crime rate.

```{r highest correlation}
data <- communities %>%
  select(-c(state, county, community, communityname, fold))
colSums(is.na(data)) %>% sort(decreasing = TRUE)
data_clean <- data %>% select(where(~ mean(is.na(.)) < 0.3))
data_clean <- data_clean %>% mutate(across(everything(), ~ replace_na(., mean(., na.rm = TRUE))))
correlations <- data_clean %>%
  select(-ViolentCrimesPerPop) %>%
  summarise(across(everything(), ~ cor(., data_clean$ViolentCrimesPerPop, use = "complete.obs"))) %>%
  pivot_longer(cols = everything(), names_to = "Variable", values_to = "Correlation") %>%
  arrange(desc(abs(Correlation)))
correlations %>%
  top_n(15, abs(Correlation)) %>%
  ggplot(aes(x = reorder(Variable, Correlation), y = Correlation, fill = Correlation > 0)) +
  geom_col(show.legend = FALSE) +
  coord_flip() +
  labs(title = "Top 15 Features Most Correlated with Violent Crime",
       x = "Variable", y = "Correlation")
```

# Adjusted r-squared

These are the top 5 variables that produce the highest r-squared score.

This is how we discovered the top 5 variables.

These are the scatterplot/linear model graphs of the top 5 variables.

```{r adjusted r-squared graphs}
```

## Possible explanations

These are some possible explanations for why certain variables are more or elss correlated with the violent crime rate.

Examples: Historical racism, racial discrepencies, slavery, and inequality, poverty and desperation, reasons for immigration, privilege

# Note that the `echo = FALSE` parameter can be added to the code chunk to prevent printing of the R code that generated the plot.