---
title: "Final Report"
author: "4677"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(dplyr)
library(tidyr)
library(corrplot)
library(caret)      
library(randomForest) 
library(plotly)
```

## Introduction

Understanding the social, economic, and demographic factors that influence crime rates has long been a central objective in the fields of criminology, sociology, and public policy.
The growing availability of detailed community-level datasets has enabled data-driven approaches to more effectively uncover the complex relationships between population characteristics and crime levels. A prominent resource in this domain is the "Communities and Crime" dataset, which aggregates data from multiple U.S. government sources and is publicly accessible through the University of California, Irvine's Machine Learning Repository.This is the original data set.

## Data description

The dataset is found at https://archive.ics.uci.edu/dataset/183/communities+and+crime and originialy created by Michael Redmond at UC Irvine in 2009, using data from 1990 and 1995. Its primary purpose is to facilitate the development of predictive models for violent crime rates—exemplified by the target variable ViolentCrimesPerPop—while enabling comprehensive analysis of the associations between community attributes and criminal outcomes.

The dataset contains 1,994 observations representing individual communities in the United States. Each observation describes a community using 122 predictive attributes and one target variable, totaling 127 fields.

The dataset documentation indicates that, apart from the first five variables—State, County, Community, CommunityName, and Fold—all of the remaining 122 variables are numeric and have been normalized to the interval [0.00, 1.00] using an equal-interval binning method. Specifically, for each variable, values exceeding three standard deviations above the mean were capped at 1.00, while those falling more than three standard deviations below the mean were set to 0.00. For example, a normalized value of 0.7 for medianIncome cannot be compared directly to a normalized value of 0.7 for PctKids2Par. Each attribute’s range is scaled independently.

Another important characteristic is the presence of missing values, especially in variables extracted from policing data. Although missing values are not abundant compared to the size of the dataset, they must still be accounted for to avoid bias during training.

The data set then lists variables to do with populations, household size, race, ages, the size of urban populations, incomes, retirements, education, employment and field of employment, divorce, families, immigration, language, home size, home ownership, rent, homeless populations, and police statistics, ending with the variable measuring the amount of violent crime proportional to the population.


```{r data set}
communities <- read_csv("data/communities.data")
head(communities)
```

## Research Question

This project is guided by three primary objectives:

1. Identify the most influential predictors of violent crime.  
Analyze demographic, socioeconomic, and law enforcement variables to determine those exhibiting the strongest statistical associations with violent crime rates.

2. Develop a predictive model for the violent crime rate.  
Train and evaluate multiple regression models using a range of machine learning techniques—including linear regression and random forests boosting—to predict the target variable ViolentCrimesPerPop. Model performance will be assessed using standard evaluation metrics such as R².

3. Ensure responsible interpretation of results.  
Given that several variables pertain to sensitive demographic attributes, including race, findings must be interpreted with caution and contextual awareness. It is critical to emphasize that correlation does not imply causation, and demographic indicators often serve as proxies for broader structural and systemic inequities rather than direct causes of criminal behavior.

## Methodology

###  1.Data Preprocessing

Several steps were taken to prepare the dataset for analysis:
```{r}
# maybe we can add some codes or not (if not just delete it) -Judy
options(repos = c(CRAN = "https://cloud.r-project.org"))

# Load column names
names_url <- "https://archive.ics.uci.edu/ml/machine-learning-databases/communities/communities.names"
name_lines <- readLines(names_url)
name_lines <- name_lines[grepl("^@attribute", name_lines)]
col_names <- gsub("^@attribute\\s+", "", name_lines)
col_names <- sub("\\s+.*$", "", col_names)
col_names <- col_names[col_names != ""]

# Load dataset
data_url <- "https://archive.ics.uci.edu/ml/machine-learning-databases/communities/communities.data"
crime <- read.csv(data_url, header = FALSE, na.strings = "?", col.names = col_names)

```

#### 1.1Handling Missing Values

Missing values were treated using:

mean or median imputation for numerical variables.

removal of attributes with a disproportionately large number of missing values.
```{r}
# the same, if we don't need it, just delete. But I think we need this step.
crime <- crime %>%
  select(-state, -county, -community, -communityname, -fold)

data_clean <- crime %>%
  mutate(across(where(is.numeric),
                ~ ifelse(is.na(.), mean(., na.rm = TRUE), .)))
cat("Data cleaned. Shape:", nrow(data_clean), "rows and", ncol(data_clean), "columns\n")
```

The dataset overall contains sparse missingness, so imputation was sufficient without significantly altering the data distribution.

#### 1.2 Converted variables to correct data types
```{r}
# if possible
```


#### 1.3 Capping procedure
After this progress, the range of the remaining values was divided into 101 equally spaced bins. Each observation was then assigned a bin number corresponding to the interval into which its value fell. For instance, in the case of the Population variable, after adjusting extreme values, the population range was partitioned into 101 bins, with each community assigned to one of these bins, as illustrated in the graph below.

```{r population visualisation, echo=FALSE}
data_clean %>% 
  ggplot(aes(x=population))+
  geom_bar()+
  labs(x="Population bin number", y="Number of communities in bin")

```


So the data set we were working with was the one containing all of these normalized values. Luckily, this method still preserves the distribution and skew of the real observations, as we can see from the plot above, most of the communities fall close to the 0.00 bin, which lines up with expectations as most communities are small. The plot also gives a good example of the treatment of extreme values, as there is a small bump at the 1.00 bin, which is caused by the few number of cities in the US with huge populations, such as New York city, Los Angeles and such.

One example of an issue that arises from the use of normalized variables comes from when we attempted to measure homeless populations. We wanted to combine the variables NumInShelters and NumStreet; however, since they were normalized, we could not combine them to compare the population to the ViolentCrimesPerPop variable. If the variables had not been normalized, this would have been possible.

There are variables that appear similar or the same but have different values. For example, BlackPerCap and RacePctBlack, have different values despite the fact that they are both normalized and measuring the same thing. In the first row, with community Lakewoodcity, RacePctBlack has a value of 0.02, while BlackPerCap has a value of 0.32. We have yet to construct a hypothesis as to why this might be true.

This is how we processed the original data set for our use.
We cleaned the original data set by setting the names to make sure that they were standardized, and converting different columns into different types. State and County, for example, was converted into factors in order to allow communities to be grouped by location if necessary.

```{r data setup}
communities <- setNames(communities, c("State","County","Community","CommunityName","Fold","Population","HouseholdSize","RacePctBlack","RacePctWhite","RacePctAsian","RacePctHisp","AgePct12t21","AgePct12t29","AgePct16t24","AgePct65up","NumbUrban","PctUrban","MedIncome","PctWWage","PctWFarmSelf","PctWInvInc","PctWSocSec","PctWPubAsst","PctWRetire","MedFamInc","PerCapInc","WhitePerCap","BlackPerCap","IndianPerCap","AsianPerCap","OtherPerCap","HispPerCap","NumUnderPov","PctPopUnderPov","PctLess9thGrade","PctNotHSGrad","PctBSorMore","PctUnemployed","PctEmploy","PctEmplManu","PctEmplProfServ","PctOccupManu","PctOccupMgmtProf","MalePctDivorce","MalePctNevMarr","FemalePctDiv","TotalPctDiv","PersPerFam","PctFam2Par","PctKids2Par","PctYoungKids2Par","PctTeen2Par","PctWorkMomYoungKids","PctWorkMom","NumIlleg","PctIlleg","NumImmig","PctImmigRecent","PctImmigRec5","PctImmigRec8","PctImmigRec10","PctRecentImmig","PctRecImmig5","PctRecImmig8","PctRecImmig10","PctSpeakEnglOnly","PctNotSpeakEnglWell","PctLargHouseFam","PctLargHouseOccup","PersPerOccupHous","PersPerOwnOccHous","PersPerRentOccHous","PctPersOwnOccup","PctPersDenseHous","PctHousLess3BR","MedNumBR","HousVacant","PctHousOccup","PctHousOwnOcc","PctVacantBoarded","PctVacMore6Mos","MedYrHousBuilt","PctHousNoPhone","PctWOFullPlumb","OwnOccLowQuart","OwnOccMedVal","OwnOccHiQuart","RentLowQ","RentMedian","RentHighQ","MedRent","MedRentPctHousInc","MedOwnCostPctInc","MedOwnCostPctIncNoMtg","NumInShelters","NumStreet","PctForeignBorn","PctBornSameState","PctSameHouse85","PctSameCity85","PctSameState85","LemasSwornFT","LemasSwFTPerPop","LemasSwFTFieldOps","LemasSwFTFieldPerPop","LemasTotalReq","LemasTotReqPerPop","PolicRepPerOffic","PolicPerPop","RacialMatchCommPol","PctPolicWhite","PctPolicBlack","PctPolicHisp","PctPolicAsian","PctPolicMinor","OfficAssgnDrugUnits","NumKindsDrugSeiz","PolicAveOTWorked","LandArea","PopDens","PctUsePubTrans","PolicCars","PolicOperBudg","LemasPctPolicOnPatr","LemasGangUnitDeploy","LemasPctOfficDrugUn","PolicBudgPerPop","ViolentCrimesPerPop"))
communities <- transform(communities,
                         State == as.factor(State),
                         County == as.factor(County),
                         Community == as.integer(Community))
```
### 2. Exploratory Data Analysis
EDA focused on understanding patterns across demographic, socio-economic, and policing domains.

```{r}
# maybe we need some summaty stats like mean median ....
```


####  2.1 Identifying the Highest Correlated Predictors
2.1.1 Purpose
The first target before constructing any predictive models was to get an understanding of community characteristics that had the strongest linear relationships with violent crime. In addition to identifying promising predictors, it also provides insight into the social patterns.

2.1.2 Method
I calculated Pearson correlation coefficients between every predictor in the dataset and the target variable ViolentCrimesPerPop. Missing values were handled using use = "complete.obs".

From these results, I extracted the top 15 variables with the strongest positive or negative correlations and visualized them using a bar plot.

2.1.3 Findings
Top15 Features Most Correlated with Violent Crime:
- Pctllleg: percentage of kids born to never married (numeric - decimal)
- racepctblack: percentage of population that is african american (numeric - decimal)
- pctWPubAsst: percentage of households with public assistance income in 1989 (numeric - decimal)
- FemalePctDiv: percentage of females who are divorced (numeric - decimal)
- TotalPctDIv: percentage of population who are divorced (numeric - decimal)
- MalePctDivorce: percentage of males who are divorced (numeric - decimal)
- PctPopUnderPov: percentage of people under the poverty level (numeric - decimal)
- PctUnemployed: percentage of people 16 and over, in the labor force, and unemployed (numeric - decimal)
- PctPersOwnOccup: percent of people in owner occupied households (numeric - decimal)
- pctWInvInc: percentage of households with investment / rent income in 1989 (numeric - decimal)
- PctTeen2Par: percent of kids age 12-17 in two parent households (numeric - decimal)
- PctYoungKids2Par: percent of kids 4 and under in two parent households (numeric - decimal)
- racePctWhite: percentage of population that is caucasian (numeric - decimal)
- PctFam2Par: percentage of families (with kids) that are headed by two parents (numeric - decimal)
- PctKids2Par: percentage of kids in family housing with two parents (numeric - decimal)

2.1.4 Interpretation: What the Top 15 Correlated Variables Reveal
The 15 variables most strongly correlated with ViolentCrimesPerPop reveal a consistent and interpretable social pattern. These variables fall into several meaningful categories that together describe the structural conditions of a community.

(a) Family Structure Variables (Very Strong Negative Correlation)
Examples include:
- PctTeen2Par — % of teens living with two parents
- PctKids2Par — % of children living with two parents
- PctYoungKids2Par — % of young children living with two parents
- PctFam2Par — % of families with two parents

Interpretation:
1) Family stability may act as a protective factor against community-level violence.
Two-parent households often correlate with:
- greater household supervision
- higher economic stability
- more time and resources to support youth
- lower likelihood of family disruption

2) These variables appear repeatedly in the top rankings because they measure similar underlying phenomena:
family cohesion, stability, and social support.

3) The strength of the correlations is especially notable because it appears across multiple child age groups—infants, kids, and teens.
This indicates that the relationship is not age-dependent, but instead a broad community structure effect.


(b) Socioeconomic Disadvantage Variables
Examples:
- PctPopUnderPov — % of population under poverty line
- PctUnemployed — unemployment rate
- pctWPubAsst — % households receiving public assistance
- PctIlleg — % of births to unmarried mothers

Interpretation:
1) Economic deprivation is a key structural factor linked with higher violent crime rates.
Communities with more poverty, unemployment, or reliance on public assistance often experience:
- limited access to resources
- fewer employment opportunities
- reduced public services
- greater stress and instability

2) PctIlleg (non-marital births) captures family and economic instability.
It is not a moral variable—it is a proxy for:
- inconsistent parental support
- unstable financial environments
- limited childcare and family resources

3) These socioeconomic indicators reinforce a classic sociological pattern:
Economic disadvantage increases social strain, which correlates with higher violent crime rates.


(c) Demographic Composition Variables (Correlation, Not Causation)
Examples:
- racepctblack
- racePctWhite

Interpretation:
These variables do show correlation but should never be interpreted as causal.

Instead, they reflect:
- the spatial concentration of racial groups due to historical housing policies
- segregation patterns
- unequal access to jobs, schools, and wealth
- unemployment and poverty burdens disproportionately affecting some communities

The correlations are capturing structural inequalities, not characteristics of individuals.


(d) Housing and Wealth Indicators
Examples:
- PctPersOwnOccup — % living in owner-occupied homes
- pctWInvInc — % with investment income

Interpretation:
Higher homeownership and investment income correlate negatively with violent crime.

This likely reflects:
- long-term community stability
- stronger neighborhood ties
- more consistent property maintenance
- greater economic buffer against shocks
- higher collective efficacy (neighbors intervene more)

These variables do not measure wealth directly but capture community-level economic security.

In general, the highest correlations indicate a coherent trend:
Violent crime closely correlates to economic distress, social unrest, and weaker family/community ties.
On the flip-side, areas with stable families, better economic and financial foundations, and long-term residents tend to have a lower incidence of violent crime.
This information helps identify which variables should be examined in greater depth.
```{r highest correlation}
communities <- read_csv("data/communities.data")
glimpse(communities)

data <- communities %>%
  select(-c(state, county, community, communityname, fold))
colSums(is.na(data)) %>% sort(decreasing = TRUE)
data_clean <- data %>% select(where(~ mean(is.na(.)) < 0.3))
data_clean <- data_clean %>% mutate(across(everything(), ~ replace_na(., mean(., na.rm = TRUE))))

str(data_clean)
summary(data_clean$ViolentCrimesPerPop)

correlations <- data_clean %>%
  select(-ViolentCrimesPerPop) %>%
  summarise(across(everything(), ~ cor(., data_clean$ViolentCrimesPerPop, use = "complete.obs"))) %>%
  pivot_longer(cols = everything(), names_to = "Variable", values_to = "Correlation") %>%
  arrange(desc(abs(Correlation)))
head(correlations, 20)

correlations %>%
  top_n(15, abs(Correlation)) %>%
  ggplot(aes(x = reorder(Variable, Correlation), y = Correlation, fill = Correlation > 0)) +
  geom_col(show.legend = FALSE) +
  coord_flip() +
  labs(title = "Top 15 Features Most Correlated with Violent Crime",
       x = "Variable", y = "Correlation")
```

#### 2.2 Adjusted R-Squared: Identifying the Most Useful Predictors
2.2.1 Purpose
While many variables exhibit a strong correlation with violent crime, most also exhibit a strong correlation with one another, particularly those pertaining to family structure. This results in multicollinearity, which tends to introduce noise to a regression model.

In order to identify the most informative predictors, I attempted a leave-one-variable-out adjusted R² approach:
Each of the 15 variables was taken away from the model, and adjusted R² was measured each time.

If adjusted R² increased when removing a variable, then that variable was redundant or detracting from the model.
If adjusted R² decreased when removing a variable, then that variable was informative.

2.2.2 Method
First, I fit a full model with all top 15 predictors. Then, for each of the 15 predictors:
- remove the variable
- re-fit the model
- compute new adjusted R²
- compare it to the full model

Finally, I sorted the results to find which removal gave the highest adjusted R².

2.2.3 Findings
Top 5 variables (adjusted R²):
- PctTeen2Par: percent of kids age 12-17 in two-parent households (numeric - decimal)
- PctPersOwnOccup: percent of people in owner-occupied households (numeric - decimal)
- racepctblack: percentage of population that is African American (numeric - decimal)
- PctUnemployed: percentage of people 16 and over, in the labor force, and unemployed (numeric - decimal)
- PctYoungKids2Par: percent of kids 4 and under in two-parent households (numeric - decimal)

2.2.4 Interpretation: Why These Variables Are the Most Important Predictors
The adjusted R² analysis identifies the variables whose removal most reduces model performance. These are the predictors that contribute the most unique explanatory power.

(a) PctTeen2Par — Most Important Predictor
Why it matters:
1) Teen-related family structure may be the most direct reflection of community environment for at-risk youth.

2) Teens are more susceptible to:
- peer pressure
- gang involvement
- unsupervised activities
- school disengagement

3) A higher percentage of teens living with two parents likely signals:
- stronger monitoring
- more consistent support
- more household resources
	
4) It outperforms similar variables (like PctKids2Par) because teenagers are more directly connected to the age groups typically involved in violent crime—either as offenders or victims.

Thus, PctTeen2Par is not just correlated; it is uniquely predictive.


(b) PctPersOwnOccup — A Key Indicator of Neighborhood Stability
Why it contributes so much predictive power:
Communities with higher homeownership generally exhibit:
- stronger long-term residency
-	stable neighborhood relationships
-	investment in local institutions
-	higher collective efficacy (neighbors watch out for each other)
-	more consistent social control mechanisms

This variable captures community qualities that are not duplicated by any single poverty or family structure variable.
It is therefore a structural stability index.


(c) racepctblack — A Statistically Important but Structural Proxy
Although the variable is statistically important, it is critical to interpret it correctly:
1) It likely captures structural inequality, not inherent traits of racial groups.
2) Because of historic redlining, wage gaps, and segregation, this variable may be acting as a proxy for:
- concentrated poverty
-	systemic underinvestment
-	school quality disparities
-	housing discrimination
-	unemployment clustering

Importantly, its statistical significance does not imply racial causation.
It reflects structural conditions disproportionately affecting certain communities.


(d) PctUnemployed — A Direct Measure of Economic Strain
Why it adds unique predictive value:
Even though poverty-related variables correlate with crime, PctUnemployed captures a more acute form of economic instability:
-	lack of steady income
-	reduced social mobility
-	higher stress and tension
-	higher idle time (especially for younger adults)

Unemployment often predicts:
-	weakened social networks
-	increased reliance on informal or illegal economies
Because unemployment reflects current economic stress, it adds information that poverty alone cannot capture.


(e) PctYoungKids2Par — Reflects Long-Term Family Stability
This variable extends the predictive strength of PctTeen2Par by showing that:
1) Family structure disadvantages start long before adolescence.
2) Communities with fewer two-parent households for young children may face long-term cumulative disadvantages, such as:
- inconsistent childcare
-	higher stress for single parents
-	limited early childhood resources
-	developmental risks that extend into later life
Although it overlaps with other family-structure variables, it still contributes enough independent signal to be selected by the adjusted R² method.


In general, the adjusted R² results enhance the correlation results by identifying which variables are significant versus being simply related.

The pattern demonstrates that stable home life, economic stability, and continuity of residence are the only unique variables that explain variance in violent crime.  Thus, demographic and poverty-related variables are still significant but become negligible when paired with the strongest predictors.


```{r adjusted r-squared}
top15_vars <- correlations %>%
  top_n(15, abs(Correlation)) %>%
  pull(Variable)
formula_top15 <- as.formula(
  paste("ViolentCrimesPerPop ~", paste(top15_vars, collapse = " + "))
)
model_top15 <- lm(formula_top15, data = communities)
summary(model_top15)

formula_full <- as.formula(
  paste("ViolentCrimesPerPop ~", paste(top15_vars, collapse = " + "))
)
model_full <- lm(formula_full, data = communities)

full_adj_r2 <- summary(model_full)$adj.r.squared 
full_adj_r2

results <- data.frame(
  Removed_Variable = character(),
  Adjusted_R2 = numeric()
)

for (v in top15_vars) {
  remaining <- top15_vars[top15_vars != v]
  f <- reformulate(remaining, "ViolentCrimesPerPop") 

  model <- lm(f, data = communities)
  results <- rbind(results, data.frame(
    Removed_Variable = v,
    Adjusted_R2 = summary(model)$adj.r.squared
  ))
}

results %>% arrange(desc(Adjusted_R2))
ggplot(results, aes(x = reorder(Removed_Variable, Adjusted_R2), y = Adjusted_R2)) +
  geom_col() +
  coord_flip() +
  labs(title = "Adjusted R² After Removing Each Variable")
```
### 3 Model Selection and Evaluation
Several regression techniques were evaluated:

#### 3.1 Linear Regression

Linear regression was used as one of our first models because it is easy to interpret and gives a good baseline for comparison. This model tries to fit a straight-line relationship between all the predictors and the violent crime rate. In other words, it assumes that each variable has a linear effect on the target. One advantage of linear regression is that we can clearly see how each predictor influences the crime rate through the model coefficients. However, the Communities and Crime dataset contains many variables that are highly correlated with each other, which can make the linear model unstable. It also struggles when the relationships in the data are more complex or non-linear. Despite these limitations, linear regression is useful for understanding the general direction of relationships and serves as a starting point before applying more advanced models.

```{r}
# this part still need some other things 
```


```{r}
library(caret)      
library(randomForest) 
set.seed(123)
train_index <- createDataPartition(data_clean$ViolentCrimesPerPop, p = 0.8, list = FALSE)
train <- data_clean[train_index, ]
test  <- data_clean[-train_index, ]
lm_model <- lm(ViolentCrimesPerPop ~ ., data = train)
lm_pred <- predict(lm_model, newdata = test)
```

#### 3.2 Tree-Based Models

```{r}
rf_model <- randomForest(ViolentCrimesPerPop ~ ., data = train, ntree = 150)
rf_pred <- predict(rf_model, newdata = test)

```

Random Forest was used as a more powerful model that can capture complex patterns. Unlike linear regression, Random Forest does not assume any simple or straight-line relationships. Instead, it builds many decision trees and combines their results, which helps improve accuracy and reduces overfitting. This makes Random Forest especially helpful for datasets like this one, where there are many predictors and the relationships between them may be non-linear. Another advantage is that Random Forest provides a ranking of feature importance, helping us see which variables contribute most to predicting violent crime. Overall, it performed better than the linear regression model and handled the complexity of the dataset more effectively.

3.2.1 Visualising Random Forest Predictions

From the plot, we see that most points follow the general upward trend, meaning the Random Forest model is able to capture the overall pattern in the data. Communities with higher actual crime rates also tend to have higher predicted values, which shows that the model is learning meaningful relationships.

However, the points are still spread out around the line, especially for larger crime values. This indicates that while the Random Forest model performs better than linear regression, it still struggles with perfect accuracy—particularly in cases with very high crime rates. This is also reflected in the RMSE value: although Random Forest lowers the prediction error, there is still some variability it cannot fully explain.

Overall, this visualisation helps confirm that the Random Forest model provides a reasonably strong fit, capturing the main pattern in the data while reducing prediction errors compared to simpler models.
```{r}
results <- data.frame(
  Actual = test$ViolentCrimesPerPop,
  Predicted = rf_pred
)
ggplot(results, aes(x = Actual, y = Predicted)) +
  geom_point(color = "blue", alpha = 0.5) +
  geom_abline(slope = 1, intercept = 0, color = "red", linetype = "dashed") +
  labs(title = "Random Forest Predictions vs Actual",
       x = "Actual Violent Crime Rate",
       y = "Predicted Crime Rate")
```

3.2.2 Residual Analysis (Model Diagnostics)

The histogram above shows the distribution of residuals from the Random Forest model. A residual is the difference between the actual violent crime rate and the predicted value. In a good predictive model, we want the residuals to be mostly close to zero, because that means the predictions are accurate.

From the plot, we can see that:

Most of the residuals are centered around 0, which means the Random Forest model generally predicts crime rates fairly well.

The distribution is roughly bell-shaped, showing that the model does not consistently over-predict or under-predict.

There are a few larger positive and negative residuals, which represent communities where the model made larger prediction errors. These outliers suggest that some areas might have unique conditions that the model cannot fully capture.

Overall, this residual plot supports that the Random Forest model performs reasonably well, with most prediction errors being small. It also explains why the Random Forest achieves a lower RMSE compared to linear regression: its residuals are more tightly grouped around zero.
```{r}

residuals_rf <- test$ViolentCrimesPerPop - rf_pred

ggplot(data.frame(residuals_rf), aes(x = residuals_rf)) +
  geom_histogram(bins = 30, fill = "darkorange", color = "white") +
  labs(title = "Residual Distribution (Random Forest)",
       x = "Residuals", y = "Count")

```

#### 3.3 Model Evaluation

R-square (R²) measures how much of the variation in violent crime can be explained by the model. A higher R-square means the model fits the data better. For example, if a model has an R-square of 0.60, it means the model explains 60% of the variation in crime rates. Linear regression usually gives a moderate R-square for this dataset, while Random Forest tends to give a higher value because it can capture more complex patterns.

RMSE (Root Mean Squared Error) measures how far the model’s predictions are from the actual crime values on average. A lower RMSE means the predictions are more accurate. RMSE is helpful because it is in the same scale as the target variable, making it easier to understand the size of the prediction errors. In our project, Random Forest produced a lower RMSE than linear regression, showing that it predicted violent crime more accurately.

Together, R-square and RMSE help us understand not only how well the model fits the data, but also how accurate the predictions are.

```{r}
lm_r2 <- cor(lm_pred, test$ViolentCrimesPerPop)^2
rf_r2 <- cor(rf_pred, test$ViolentCrimesPerPop)^2

lm_rmse <- sqrt(mean((lm_pred - test$ViolentCrimesPerPop)^2))
rf_rmse <- sqrt(mean((rf_pred - test$ViolentCrimesPerPop)^2))

cat("Linear Regression: R2 =", lm_r2, " RMSE =", lm_rmse, "\n")
cat("Random Forest: R2 =", rf_r2, " RMSE =", rf_rmse, "\n")

```

### 4.Improvements and Ethical Considerations
Possible Improvements

Feature selection to reduce noise

Hyperparameter tuning for deeper trees or more estimators

Try Gradient Boosting or XGBoost

Scale features for linear regression to improve stability

Ethical Concerns

Working with crime data raises several ethical issues:

Bias amplification
Models trained on biased data may reinforce unfair stereotypes about communities or racial groups.

Misuse in policing
Predictive models could be used for over-policing certain neighbourhoods.

Privacy
Socio-economic data must be handled responsibly.
It is crucial to treat predictions carefully and avoid interpreting them as definitive or causal.

### 5. Results & feedback

In general, the models performed as follows:

Linear Regression: Moderate accuracy; struggled with multicollinearity.

Random Forest: Good performance; provided useful feature-importance rankings.

These results suggest that the relationship between community attributes and violent crime is highly non-linear, making tree-based ensemble methods particularly suitable.

### 6. Overall Summary

This project explored the Communities and Crime dataset to investigate whether socio-economic variables can predict violent crime rates.

Key Findings

Violent crime correlates with poverty, unemployment, and family structure.

Linear Regression provided moderate accuracy (R² ≈ 0.48).

Random Forest performed strongly (R² ≈ 0.80) with much lower error.

Visualisations confirmed that Random Forest made significantly better predictions.

Ethical concerns must be considered when working with sensitive social data.

Conclusion

Random Forest is a highly effective method for predicting violent crime rates using community-level features.
The dataset highlights the complex relationships between crime and socio-economic conditions, and machine learning models can help uncover these patterns—provided they are used responsibly.


## Possible explanations

There are many possible explanations for why certain variables are more or less correlated with the violent crime rate. 
PctIlleg likely has a high correlation due to the lack of legal work opportunities for people who are not in the United States on a Visa, Green Card, or who are citizens, due to many employeres requiring things like SSNs and proof of visas. As such, many illegal immigrants may be forced to turn to crime, creating a larger crime community, which can lead to an increased prevalence of violence due to increases in power.
Other things like historical racism, slavery, bigotry, inequality, poverty and desperation, reasons for immigration, and the distribution of certain types of privilege are all related to why certain variables have high or low correlations with the prevalence of violent crime.

Note that the `echo = FALSE` parameter can be added to the code chunk to prevent printing of the R code that generated the plot.