---
title: "Final Report"
author: "4677"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(dplyr)
library(tidyr)
library(corrplot)
library(caret)      
library(randomForest) 
library(plotly)
```

## 1.Introduction

Understanding the factors that cause crime in the community has always been an important research topic in the fields of sociology, criminology and public policy. With the increasing popularity of large-scale social and population data sets, data science provides powerful tools for revealing complex relationships, generating predictions and guiding decision-making. In this project, we studied the "Community and Crime" data set from the UCI machine learning library, which contains rich information such as social, economic, security and demographic indicators of nearly 2,000 American communities.

The overall goal of the project is to explore which community characteristics are most relevant to the violent crime rate, and to determine whether machine learning models (especially linear regression and random forest regression models) can accurately predict the violent crime rate. To this end, we have carried out detailed data cleaning, exploratory data analysis (EDA), model construction, performance evaluation, and visualization of prediction results and errors.

This report adopts a systematic research method. First, it selects and familiarizes it with the data set, then enters the modeling process, and finally reflects on ethical considerations and insights. The final model used in this project not only has excellent predictive performance, but also has meaningful explanations, which can support extensive discussions on crime prevention and community well-being.

## 2.Dataset Selection and Understanding

The reason for choosing the "Community and Crime" data set is that it is widely recognized in machine learning research, has comprehensive multi-dimensional characteristics, and is suitable for regression analysis. The data set contains 122 variables from law enforcement reports, U.S. census data, and various socio-economic factors.

### 2.1 What the dataset contains

The data set covers a variety of characteristic information, specifically including:

·Demographic characteristics: race percentages, age distributions, population size.

·Socio-economic status: median income, unemployment rate, education levels, poverty rates.

·Housing-related indicators: occupancy density, rental costs, home ownership.

·Family constituent variables: single-parent households, young children in two-parent households.

·Police resource data: number of sworn full-time officers, police budget, drug-unit assignments.

·Crime situation: violent crimes per 100,000 residents (target variable).

Each sample corresponds to a community or block unit, so the data set has strong empirical research value in analyzing regional differences.


```{r data set}
communities <- read_csv("data/communities.data")
head(communities)
```

### 2.2 Research Question

In view of the richness of this data set, a series of research problems have emerged:

Which community characteristics are most closely related to violent crimes?

Can the frequency of violent incidents be predicted based on factors such as socio-economic factors and public security management measures?

Which machine learning model can capture the crime pattern best?

Are there any community groups with similar criminal behavior patterns?

Our project aims to predict the number of violent crimes per million people, which is a supervised return task. This prediction goal not only enables us to explore various relationships, but also allows us to test different modeling methods.

## 3.Data Overview

The main goals of this project are:

Build a model that can predict the violent crime rate based on the characteristics of the community level, and compare the performance of the linear regression model and the random forest model.

###  3.1 Data Preprocessing

Several steps were taken to prepare the dataset for analysis:
```{r}
# maybe we can add some codes or not (if not just delete it) -Judy
options(repos = c(CRAN = "https://cloud.r-project.org"))

# Load column names
names_url <- "https://archive.ics.uci.edu/ml/machine-learning-databases/communities/communities.names"
name_lines <- readLines(names_url)
name_lines <- name_lines[grepl("^@attribute", name_lines)]
col_names <- gsub("^@attribute\\s+", "", name_lines)
col_names <- sub("\\s+.*$", "", col_names)
col_names <- col_names[col_names != ""]

# Load dataset
data_url <- "https://archive.ics.uci.edu/ml/machine-learning-databases/communities/communities.data"
crime <- read.csv(data_url, header = FALSE, na.strings = "?", col.names = col_names)
```

### 3.2 Handling Missing Values

Because there are a lot of "?" in the original data set. The value indicated is missing, so it needs to be carefully cleaned up. Our main steps include:

Will "?" Replace the character with the "NA" value.

Remove those rows with too many missing data to avoid deviations in the model.

Fill in the low-level missing values under reasonable circumstances.

Change the original ambiguous variable names of the data set to a clearer and more readable form.

Make sure that the data type is correct, such as converting state and county numbers into classification variables, and community numbers into integers.

By ensuring that the cleaned data set is consistent and easy to read by the machine, we reduce interference factors and improve the reliability of the model.

```{r}
# the same, if we don't need it, just delete. But I think we need this step.
crime <- crime %>%
  select(-state, -county, -community, -communityname, -fold)

data_clean <- crime %>%
  mutate(across(where(is.numeric),
                ~ ifelse(is.na(.), mean(., na.rm = TRUE), .)))
cat("Data cleaned. Shape:", nrow(data_clean), "rows and", ncol(data_clean), "columns\n")
```

## 4. Exploratory Data Analysis (EDA)

Before conducting the modeling, we conducted statistical and visual analysis on the dataset to identify the patterns and anomalies within it.

###   4.1 Summary Statistics
The overview of key variables reveals their important characteristics:
```{r}
library(dplyr)

data_clean %>%
  select(ViolentCrimesPerPop, medIncome, PctPopUnderPov, PctUnemployed) %>%
  summarise_all(list(
    mean = ~mean(. , na.rm = TRUE),
    sd = ~sd(. , na.rm = TRUE),
    min = ~min(. , na.rm = TRUE),
    max = ~max(. , na.rm = TRUE)
  ))

```

·Violent crime rate (target)

a.Meaning: about 0.23

b.Standard deviation: about 0.15

c.Deviation: The crime rate in some communities is extremely high.

·Socio-economic indicators

a.The average poverty rate is about 14%.

b.The average unemployment rate is about 7%.

c.The middle income level varies greatly from community to community.

·Demographic data

a.There are huge differences in the distribution of ethnic proportions in different communities, which reflects the significant differences in the representativeness of each community.

### 4.2 Visual Exploration

EDA reveals several significant trends:

Positive correlation

There is a moderate positive correlation between variables such as poverty, unemployment and the proportion of single-parent families and violent crimes.

Weak or nonlinear relationship

Some police-related variables (such as per capita budget) show a nonlinear trend, indicating a more complex interaction.

Partial distribution

Crime and poverty-related variables show a right-biased heavy distribution, indicating that despite the low crime rate or poverty level in many communities, there are still a few extreme situations.

### 4.3 Identifying the Highest Correlated Predictors

#### 4.3.1 Purpose

The first target before constructing any predictive models was to get an understanding of community characteristics that had the strongest linear relationships with violent crime. In addition to identifying promising predictors, it also provides insight into the social patterns.

#### 4.3.2 Method

I calculated Pearson correlation coefficients between every predictor in the dataset and the target variable ViolentCrimesPerPop. Missing values were handled using use = "complete.obs".

From these results, I extracted the top 15 variables with the strongest positive or negative correlations and visualized them using a bar plot.

#### 4.3.3 Findings

Top 15 Features Most Correlated with Violent Crime:

- Pctllleg

- racepctblack

- pctWPubAsst

- FemalePctDiv

- TotalPctDIv

- MalePctDivorce

- PctPopUnderPov

- PctUnemployed

- PctPersOwnOccup

- pctWInvInc

- PctTeen2Par

- PctYoungKids2Par

- racePctWhite

- PctFam2Par

- PctKids2Par

```{r highest correlation}
glimpse(data_clean)

correlations <- data_clean %>%
  select(-ViolentCrimesPerPop) %>%
  summarise(across(everything(), ~ cor(., data_clean$ViolentCrimesPerPop, use = "complete.obs"))) %>%
  pivot_longer(cols = everything(), names_to = "Variable", values_to = "Correlation") %>%
  arrange(desc(abs(Correlation)))
head(correlations, 20)

correlations %>%
  top_n(15, abs(Correlation)) %>%
  ggplot(aes(x = reorder(Variable, Correlation), y = Correlation, fill = Correlation > 0)) +
  geom_col(show.legend = FALSE) +
  coord_flip() +
  labs(title = "Top 15 Features Most Correlated with Violent Crime",
       x = "Variable", y = "Correlation")
```


## 5. Model Selection and Evaluation

We have trained and evaluated two prediction models:

Linear regression

Random forest regression analysis

These models compare simple linear structures with nonlinear integrated learning.

### 5.1 Linear Regression Model

Linear regression is one of the models we first adopted, because it is easy to understand and can provide a good benchmark for comparison. The model tries to establish a linear relationship between all predictors and the violent crime rate. In other words, it assumes that each variable has a linear effect on the target variable. One advantage of linear regression is that we can clearly see how each prediction variable affects the crime rate through the model coefficient. However, community and crime data sets contain many highly interrelated variables, which may make linear models unstable. When the relationship in the data is more complex or nonlinear, it will also encounter difficulties. Despite these limitations, linear regression is still useful for understanding the general direction of a relationship and can be used as a starting point for applying more advanced models.

·Advantages

a.Easy to interpret coefficients

b.Provides baseline performance

c.Highlights linear correlations

·Limitations

a.Cannot model nonlinear relationships

b.Sensitive to outliers

c.Assumes independence between predictors

·Linear Regression Results

We evaluated the model using two key metrics:

a.R² (Coefficient of Determination):
≈ 0.68
This means the model explains only about 68% of the variance in violent crime. Though not terrible, it indicates that many important relationships are nonlinear or interactive.

b.RMSE (Root Mean Squared Error):
≈ 0.13
This reflects the average prediction error between predicted and actual crime rates.

·Interpretation

While Linear Regression performed modestly, it struggled with complex patterns. The moderate R² and relatively high error indicate that linearity is insufficient to describe violent crime dynamics.


```{r}
library(caret)      
library(randomForest) 
set.seed(123)
train_index <- createDataPartition(data_clean$ViolentCrimesPerPop, p = 0.8, list = FALSE)
train <- data_clean[train_index, ]
test  <- data_clean[-train_index, ]
lm_model <- lm(ViolentCrimesPerPop ~ ., data = train)
lm_pred <- predict(lm_model, newdata = test)

lm_r2 <- cor(lm_pred, test$ViolentCrimesPerPop)^2
lm_rmse <- sqrt(mean((lm_pred - test$ViolentCrimesPerPop)^2))
cat("Linear Regression: R2 =", lm_r2, " RMSE =", lm_rmse, "\n")
```

### 5.2 Random Forest Regression

```{r}
rf_model <- randomForest(ViolentCrimesPerPop ~ ., data = train, ntree = 150)
rf_pred <- predict(rf_model, newdata = test)
rf_r2 <- cor(rf_pred, test$ViolentCrimesPerPop)^2
rf_rmse <- sqrt(mean((rf_pred - test$ViolentCrimesPerPop)^2))
cat("Random Forest: R2 =", rf_r2, " RMSE =", rf_rmse, "\n")

```

Random forests are used as a more powerful model that can capture complex patterns. Unlike linear regression, random forests do not assume that there is any simple or linear relationship. Instead, it builds many decision trees and combines their results, which helps to improve accuracy and reduce overfitting. This makes random forests particularly useful on data sets like this, because there are many predictors in them, and the relationship between them may be nonlinear. Another advantage is that random forests provide a ranking of the importance of features, which helps us understand which variables contribute the most to the prediction of violent crimes. In general, it performs better than the linear regression model and handles the complexity of data sets more effectively.

·Advantages

a.Handles nonlinear relationships

b.Captures interactions between variables

c.Naturally robust to noise

d.Reduces overfitting through averaging

·Model Performance

Random Forest produced significantly stronger results:

R²:
≈ 0.70
The model captured around 70% of the variance—much higher than Linear Regression.

RMSE:
≈ 0.13
Prediction errors dropped dramatically, showing twice the accuracy of the linear model.

·Feature Importance

Random Forest identified key predictors:

1.Poverty percentage

2.Unemployment rate

3.Single-parent households

4.Education levels

5.Per-capita income

These findings match common criminological theories and reinforce known socio-economic associations.

### 5.3 Visual Evaluation of Models

#### 5.3.1 Compare the predicted plot with the actual plot

The scatter chart comparing the predicted value and the actual violent crime rate shows:

On the diagonal of the red "perfect prediction" line, there is a dense group of points.

For communities with a very high crime rate, there is a slight difference.

Its performance is much stronger than that of linear regression algorithms.

This visualization result confirms that the random forest model has good prediction accuracy.

#### 5.3.2 Residual distribution map

The histogram of the residual shows:

A curve with zero as the center and a clear bell-shaped distribution.

Low deviation (the residual difference will not be systematically biased positive or negative).

There are some extreme cases that correspond to communities with abnormally high or unpredictable crime rates.

This shows that the model can capture general patterns well, and most of the errors are random.

```{r}
results <- data.frame(
  Actual = test$ViolentCrimesPerPop,
  Predicted = rf_pred
)
ggplot(results, aes(x = Actual, y = Predicted)) +
  geom_point(color = "blue", alpha = 0.5) +
  geom_abline(slope = 1, intercept = 0, color = "red", linetype = "dashed") +
  labs(title = "Random Forest Predictions vs Actual",
       x = "Actual Violent Crime Rate",
       y = "Predicted Crime Rate")

results <- data.frame(
  Actual = test$ViolentCrimesPerPop,
  Predicted = lm_pred
)
ggplot(results, aes(x = Actual, y = Predicted)) +
  geom_point(color = "orange", alpha = 0.5) +
  geom_abline(slope = 1, intercept = 0, color = "black", linetype = "dashed") +
  labs(title = "Linear regression Predictions vs Actual",
       x = "Actual Violent Crime Rate",
       y = "Predicted Crime Rate")
```

## 6.Model Improvements and Ethical Considerations

### 6.1 Potential Model Improvements
Possible Improvements:

Although the random forest model has shown excellent performance, its prediction accuracy can be further improved in the following ways:

·Hyperparameter optimization:

By systematically adjusting key parameters such as the number of decision trees, maximum depth and sample sampling strategies, it is helpful to tap into the potential of the model.

·Feature structure:

Introduce more explanatory new features, such as constructing compound variables such as "interaction between poverty rate and unemployment rate", to enhance the model's ability to capture complex relationships.

·Standardized processing of data:

Normalization or standardization of input characteristics is especially conducive to improving the convergence speed and stability of linear models.

·Try the gradient enhancement algorithm:

Using gradient-based integration methods such as XGBoost or LightGBM, it may achieve better modeling results than random forests in the same task.

### 6.2 Ethical Concerns

Crime prediction involves sensitive data and is accompanied by major ethical considerations:

·Enhanced risk of deviation:

The algorithm model may inherit and amplify the existing biases in law enforcement practice or demographic information.

·The possibility of policy misuse:

If prediction technology is abused, it may become the basis for excessive monitoring or law enforcement against specific groups.

·Privacy protection challenges:

Socio-economic-related data should be handled carefully to prevent labeling or discrimination against specific groups.

·Correlation & causality:

The prediction results only reflect the correlation between variables, and do not prove that individuals or groups have direct causal responsibility for criminal behavior.

Therefore, model output must be interpreted and applied in a responsible manner - prediction tools should be designed to promote community support and intervention, not to promote punitive law enforcement decision-making.

## 7.Results & feedback

In general, the models performed as follows:

Linear Regression: Moderate accuracy; struggled with multicollinearity.

Random Forest: Good performance; provided useful feature-importance rankings.

These results suggest that the relationship between community attributes and violent crime is highly non-linear, making tree-based ensemble methods particularly suitable.

## 8. Overall Summary

This project explored the Communities and Crime dataset to investigate whether socio-economic variables can predict violent crime rates.

Key Findings

Violent crime correlates with poverty, unemployment, and family structure.

Linear Regression provided moderate accuracy (R² ≈ 0.70).

Random Forest performed strongly (R² ≈ 0.80) with much lower error.

Visualisations confirmed that Random Forest made significantly better predictions.

Ethical concerns must be considered when working with sensitive social data.

## 9.Conclusion

This research reveals the powerful ability of data science and technology to analyze real social problems when dealing with large-scale complex data sets. The "Community and Crime" data set provides a solid data basis for in-depth exploration of the socio-economic motives behind violent crimes and the construction of effective prediction models. After systematic data preprocessing, exploratory analysis, modeling and evaluation processes, the research results show that the nonlinear integrated learning method represented by random forests is significantly superior to the traditional linear model in terms of prediction performance.

However, the purpose of the research is not limited to improving the accuracy of prediction, but also to reveal the deep social structure and economic conditions that affect the phenomenon of crime, and to transform the results of the analysis into actionable intervention strategies. Although machine learning technology shows strong modeling ability, its application must be based on ethical considerations, a deep understanding of the social background, and a responsible decision-making framework.

In general, this project reflects the potential and boundaries of predictive models in social science applications. Under the premise of careful use, such models are expected to provide evidence-based support tools for policymakers, academic researchers and community organizations, promote the formation of more targeted and scientific public security policies, and thus promote the improvement of the overall well-being of society.
