---
title: "Final Report"
author: "4677"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(dplyr)
library(tidyr)
library(corrplot)
library(caret)      
library(randomForest) 
library(plotly)
```

## Introduction

Understanding the social, economic, and demographic factors that influence crime rates has long been a central objective in the fields of criminology, sociology, and public policy.
The growing availability of detailed community-level datasets has enabled data-driven approaches to more effectively uncover the complex relationships between population characteristics and crime levels. A prominent resource in this domain is the "Communities and Crime" dataset, which aggregates data from multiple U.S. government sources and is publicly accessible through the University of California, Irvine's Machine Learning Repository.This is the original data set.

## Data description

The dataset is found at https://archive.ics.uci.edu/dataset/183/communities+and+crime and originialy created by Michael Redmond at UC Irvine in 2009, using data from 1990 and 1995. Its primary purpose is to facilitate the development of predictive models for violent crime rates—exemplified by the target variable ViolentCrimesPerPop—while enabling comprehensive analysis of the associations between community attributes and criminal outcomes.

The dataset contains 1,994 observations representing individual communities in the United States. Each observation describes a community using 122 predictive attributes and one target variable, totaling 127 fields.

The dataset documentation indicates that, apart from the first five variables—State, County, Community, CommunityName, and Fold—all of the remaining 122 variables are numeric and have been normalized to the interval [0.00, 1.00] using an equal-interval binning method. Specifically, for each variable, values exceeding three standard deviations above the mean were capped at 1.00, while those falling more than three standard deviations below the mean were set to 0.00. For example, a normalized value of 0.7 for medianIncome cannot be compared directly to a normalized value of 0.7 for PctKids2Par. Each attribute’s range is scaled independently.

Another important characteristic is the presence of missing values, especially in variables extracted from policing data. Although missing values are not abundant compared to the size of the dataset, they must still be accounted for to avoid bias during training.

The data set then lists variables to do with populations, household size, race, ages, the size of urban populations, incomes, retirements, education, employment and field of employment, divorce, families, immigration, language, home size, home ownership, rent, homeless populations, and police statistics, ending with the variable measuring the amount of violent crime proportional to the population.


```{r data set}
communities <- read_csv("data/communities.data")
head(communities)
```

## Research Question

This project is guided by three primary objectives:

1. Identify the most influential predictors of violent crime.  
Analyze demographic, socioeconomic, and law enforcement variables to determine those exhibiting the strongest statistical associations with violent crime rates.

2. Develop a predictive model for the violent crime rate.  
Train and evaluate multiple regression models using a range of machine learning techniques—including linear regression and random forests boosting—to predict the target variable ViolentCrimesPerPop. Model performance will be assessed using standard evaluation metrics such as R².

3. Ensure responsible interpretation of results.  
Given that several variables pertain to sensitive demographic attributes, including race, findings must be interpreted with caution and contextual awareness. It is critical to emphasize that correlation does not imply causation, and demographic indicators often serve as proxies for broader structural and systemic inequities rather than direct causes of criminal behavior.

## Methodology

####  1.Data Preprocessing

Several steps were taken to prepare the dataset for analysis:
```{r}
# maybe we can add some codes or not (if not just delete it) -Judy
options(repos = c(CRAN = "https://cloud.r-project.org"))

# Load column names
names_url <- "https://archive.ics.uci.edu/ml/machine-learning-databases/communities/communities.names"
name_lines <- readLines(names_url)
name_lines <- name_lines[grepl("^@attribute", name_lines)]
col_names <- gsub("^@attribute\\s+", "", name_lines)
col_names <- sub("\\s+.*$", "", col_names)
col_names <- col_names[col_names != ""]

# Load dataset
data_url <- "https://archive.ics.uci.edu/ml/machine-learning-databases/communities/communities.data"
crime <- read.csv(data_url, header = FALSE, na.strings = "?", col.names = col_names)

```

#### 2.Handling Missing Values

Missing values were treated using:

mean or median imputation for numerical variables.

removal of attributes with a disproportionately large number of missing values.
```{r}
# the same, if we don't need it, just delete. But I think we need this step.
crime <- crime %>%
  select(-state, -county, -community, -communityname, -fold)

data_clean <- crime %>%
  mutate(across(where(is.numeric),
                ~ ifelse(is.na(.), mean(., na.rm = TRUE), .)))
cat("Data cleaned. Shape:", nrow(data_clean), "rows and", ncol(data_clean), "columns\n")
```

The dataset overall contains sparse missingness, so imputation was sufficient without significantly altering the data distribution.


#### 3.Capping procedure
After this progress, the range of the remaining values was divided into 101 equally spaced bins. Each observation was then assigned a bin number corresponding to the interval into which its value fell. For instance, in the case of the Population variable, after adjusting extreme values, the population range was partitioned into 101 bins, with each community assigned to one of these bins, as illustrated in the graph below.

```{r population visualisation, echo=FALSE}
data_clean %>% 
  ggplot(aes(x=population))+
  geom_bar()+
  labs(x="Population bin number", y="Number of communities in bin")

```
So the data set we were working with was the one containing all of these normalized values. Luckily, this method still preserves the distribution and skew of the real observations, as we can see from the plot above, most of the communities fall close to the 0.00 bin, which lines up with expectations as most communities are small. The plot also gives a good example of the treatment of extreme values, as there is a small bump at the 1.00 bin, which is caused by the few number of cities in the US with huge populations, such as New York city, Los Angeles and such.

One example of an issue that arises from the use of normalized variables comes from when we attempted to measure homeless populations. We wanted to combine the variables NumInShelters and NumStreet; however, since they were normalized, we could not combine them to compare the population to the ViolentCrimesPerPop variable. If the variables had not been normalized, this would have been possible.

There are variables that appear similar or the same but have different values. For example, BlackPerCap and RacePctBlack, have different values despite the fact that they are both normalized and measuring the same thing. In the first row, with community Lakewoodcity, RacePctBlack has a value of 0.02, while BlackPerCap has a value of 0.32. We have yet to construct a hypothesis as to why this might be true.

This is how we processed the original data set for our use.
We cleaned the original data set by setting the names to make sure that they were standardized, and converting different columns into different types. State and County, for example, was converted into factors in order to allow communities to be grouped by location if necessary.

```{r data setup}
communities <- setNames(communities, c("State","County","Community","CommunityName","Fold","Population","HouseholdSize","RacePctBlack","RacePctWhite","RacePctAsian","RacePctHisp","AgePct12t21","AgePct12t29","AgePct16t24","AgePct65up","NumbUrban","PctUrban","MedIncome","PctWWage","PctWFarmSelf","PctWInvInc","PctWSocSec","PctWPubAsst","PctWRetire","MedFamInc","PerCapInc","WhitePerCap","BlackPerCap","IndianPerCap","AsianPerCap","OtherPerCap","HispPerCap","NumUnderPov","PctPopUnderPov","PctLess9thGrade","PctNotHSGrad","PctBSorMore","PctUnemployed","PctEmploy","PctEmplManu","PctEmplProfServ","PctOccupManu","PctOccupMgmtProf","MalePctDivorce","MalePctNevMarr","FemalePctDiv","TotalPctDiv","PersPerFam","PctFam2Par","PctKids2Par","PctYoungKids2Par","PctTeen2Par","PctWorkMomYoungKids","PctWorkMom","NumIlleg","PctIlleg","NumImmig","PctImmigRecent","PctImmigRec5","PctImmigRec8","PctImmigRec10","PctRecentImmig","PctRecImmig5","PctRecImmig8","PctRecImmig10","PctSpeakEnglOnly","PctNotSpeakEnglWell","PctLargHouseFam","PctLargHouseOccup","PersPerOccupHous","PersPerOwnOccHous","PersPerRentOccHous","PctPersOwnOccup","PctPersDenseHous","PctHousLess3BR","MedNumBR","HousVacant","PctHousOccup","PctHousOwnOcc","PctVacantBoarded","PctVacMore6Mos","MedYrHousBuilt","PctHousNoPhone","PctWOFullPlumb","OwnOccLowQuart","OwnOccMedVal","OwnOccHiQuart","RentLowQ","RentMedian","RentHighQ","MedRent","MedRentPctHousInc","MedOwnCostPctInc","MedOwnCostPctIncNoMtg","NumInShelters","NumStreet","PctForeignBorn","PctBornSameState","PctSameHouse85","PctSameCity85","PctSameState85","LemasSwornFT","LemasSwFTPerPop","LemasSwFTFieldOps","LemasSwFTFieldPerPop","LemasTotalReq","LemasTotReqPerPop","PolicRepPerOffic","PolicPerPop","RacialMatchCommPol","PctPolicWhite","PctPolicBlack","PctPolicHisp","PctPolicAsian","PctPolicMinor","OfficAssgnDrugUnits","NumKindsDrugSeiz","PolicAveOTWorked","LandArea","PopDens","PctUsePubTrans","PolicCars","PolicOperBudg","LemasPctPolicOnPatr","LemasGangUnitDeploy","LemasPctOfficDrugUn","PolicBudgPerPop","ViolentCrimesPerPop"))
communities <- transform(communities,
                         State == as.factor(State),
                         County == as.factor(County),
                         Community == as.integer(Community))
```
### 2. Exploratory Data Analysis
EDA focused on understanding patterns across demographic, socio-economic, and policing domains.

## Highest Correlation

These are the variables with the most significant correlation to the violent crime rate.

```{r highest correlation}
data <- communities %>%
  select(-c(State, County, Community, CommunityName, Fold))
colSums(is.na(data)) %>% sort(decreasing = TRUE)
data_clean <- data %>% select(where(~ mean(is.na(.)) < 0.3))
data_clean <- data_clean %>% mutate(across(everything(), ~ replace_na(., mean(., na.rm = TRUE))))
correlations <- data_clean %>%
  select(-ViolentCrimesPerPop) %>%
  summarise(across(everything(), ~ cor(., data_clean$ViolentCrimesPerPop, use = "complete.obs"))) %>%
  pivot_longer(cols = everything(), names_to = "Variable", values_to = "Correlation") %>%
  arrange(desc(abs(Correlation)))
correlations %>%
  top_n(15, abs(Correlation)) %>%
  ggplot(aes(x = reorder(Variable, Correlation), y = Correlation, fill = Correlation > 0)) +
  geom_col(show.legend = FALSE) +
  coord_flip() +
  labs(title = "Top 15 Features Most Correlated with Violent Crime",
       x = "Variable", y = "Correlation")
```

## Adjusted r-squared

These are the top 5 variables that produce the highest adjusted r-squared score.

```{r adjusted r-squared}

```

This is how we discovered the top 5 variables.

These are the scatterplot/linear model graphs of the top 5 variables.

```{r adjusted r-squared graphs, echo = FALSE}

```

## Possible explanations

There are many possible explanations for why certain variables are more or less correlated with the violent crime rate. 
PctIlleg likely has a high correlation due to the lack of legal work opportunities for people who are not in the United States on a Visa, Green Card, or who are citizens, due to many employeres requiring things like SSNs and proof of visas. As such, many illegal immigrants may be forced to turn to crime, creating a larger crime community, which can lead to an increased prevalence of violence due to increases in power.
Other things like historical racism, slavery, bigotry, inequality, poverty and desperation, reasons for immigration, and the distribution of certain types of privilege are all related to why certain variables have high or low correlations with the prevalence of violent crime.

Note that the `echo = FALSE` parameter can be added to the code chunk to prevent printing of the R code that generated the plot.